---
title: Data visualization with ggplot2
minutes: 60
---



```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
surveys_complete <- read_csv(file = "data_output/surveys_complete.csv", col_types = cols())
```

```{r, echo=FALSE, purl=TRUE}
### Data Visualization with ggplot2
```

------------

> ### Learning Objectives
>
> * Produce scatter plots, boxplots, and time series plots using ggplot.
> * Set universal plot settings.
> * Describe what faceting is and apply faceting in ggplot.
> * Modify the aesthetics of an existing ggplot plot (including axis labels and color).
> * Build complex and customized plots from data in a data frame.

--------------

We start by loading the required packages. **`ggplot2`** is included in the **`tidyverse`** package.

```{r load-tidyverse-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

Later we will learn how to read in data in R, but for now we are going to use a clean and ready to use dataset that is provided by the **`ratdat`** data package. To make this dataset available in memory, we need to load this package. We will use the `portal_dipo` dataset.


```{r load-ratdat-package,message=FALSE, purl=TRUE}
library(ratdat)
portal_dipo
```                                      


## Plotting with **`ggplot2`**

**`ggplot2`** is a powerful plotting package that makes it simple to create
complex plots from data in a data frame. It provides a programmatic interface
for specifying what variables to plot, how they are displayed, and general
visual properties. Therefore, we only need small changes to the code if the
underlying data change or if we decide to change from a bar plot to a
scatterplot. This interface helps in creating publication quality plots with
minimal amounts of adjustments and tweaking.

**`ggplot2`** functions like data in the 'long' tidy format, i.e., a column for
every dimension, and a row for every observation. Well-structured data will save
you lots of time when making figures with **`ggplot2`**. We start learning R by using **`ggplot2`** because it relies on concepts that we will need when we talk about data transformation in the next lessons.

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```r
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()
```

- use the `ggplot()` function and bind the plot to a specific data frame using the `data` argument

```{r, eval=FALSE, purl=FALSE}
ggplot(data = portal_dipo)
```

- define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.

```{r, eval=FALSE, purl=FALSE}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length))
```

- add 'geoms' – graphical representations of the data in the plot (points,
  lines, bars). **`ggplot2`** offers many different geoms; we will use some 
  common ones today, including:
  
      * `geom_point()` for scatter plots, dot plots, etc.
      * `geom_boxplot()` for, well, boxplots!
      * `geom_line()` for trend lines, time series, etc.  

To add a geom to the plot use the `+` operator. Because we have two continuous variables, let's use `geom_point()` first to create a scatter plot:

```{r first-ggplot, purl=FALSE}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows
you to modify existing `ggplot` objects. This means you can easily set up plot
templates and explore different types of plots, so the above plot can also be
generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl=FALSE}
## Assign plot to a variable
surveys_plot <- ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length))

## Draw the plot
surveys_plot + 
    geom_point()
```

## Building your plots iteratively

<!-- the instructor guide should explain that for this section, the instructor should modify the same line of code instead of retyping each iteration -->

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, purl=FALSE}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point()
```

Now, we can start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to avoid overplotting:

```{r adding-transparency, purl=FALSE}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.2)
```

It might also be a good idea to use jittering:

<!-- explain what jittering is -->


```{r jittering, purl=FALSE}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.2,  position = position_jitter())
```

We can also add colors for all the points:

```{r adding-colors, purl=FALSE}					
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1, color = "blue")
```

Or to color each species in the plot differently, you could specify the variable to use to the `color` argument within `aes()`:


```{r color-by-species, purl=FALSE}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1, aes(color = species_id))
```

In this case **`ggplot2`** will automatically assign one color for each species.

> ### Challenge
>
> 1. use of `shape` to differentiate males and females
> 2. color by year. What happens? Why is the scale different from when coloring by species? 

<!--- add short section on exploring feature of dataset, if tibble type of columns shown, otherwise need to introduce str() --->


## Faceting

**`ggplot2`** has a special technique called *faceting* that allows the user to
split one plot into multiple plots based on discrete variables included in the
dataset. We will use it to make a time series plot for each species:

```{r}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point(aes(color = species_id, shape = sex),
               position = position_jitter(), alpha = .3) +
    facet_wrap(~ species_id)
```

```{r}
ggplot(data = portal_dipo, aes(x = weight, y = hindfoot_length)) +
    geom_point(aes(color = species_id, shape = sex),
               position = position_jitter(), alpha = .3) +
    facet_grid(species_id ~ year)
```

> ### Challenge
>
> Create a visualization that uses facetting with the sex variable


## Boxplots


We can use boxplots to visualize the distribution of weight within each species:

```{r boxplot, purl=FALSE}
ggplot(data = portal_dipo, aes(x = species_id, y = weight)) +
    geom_boxplot()
```

By adding points to boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-points, purl=FALSE}
ggplot(data = portal_dipo, aes(x = species_id, y = weight)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.3, color = "tomato")
```

Notice how the boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot in front of the points such that it's not
hidden?

> ### Challenges
>
> Boxplots are useful summaries, but hide the *shape* of the distribution. For
> example, if the distribution is bimodal, we would not see it in a
> boxplot. An alternative to the boxplot is the violin plot, where the shape 
(of the density of points) is drawn.
>
> - Replace the box plot with a violin plot; see `geom_violin()`.
>
> In many types of data, it is important to consider the *scale* of the
> observations.  For example, it may be worth changing the scale of the axis to
> better distribute the observations in the space of the plot.  Changing the scale
> of the axes is done similarly to adding/modifying other components (i.e., by
> incrementally adding commands). Try making these modifications:
>
> - Represent weight on the log~10~ scale; see `scale_y_log10()`.
>
> So far, we've looked at the distribution of weight within species. Try making a
> new plot to explore the distribution of another variable within each species.
>
> - Create a boxplot for `hindfoot_length`. Overlay the boxplot layer on a jitter
>   layer to show actual measurements.

```{r boxplot-challenge, eval=FALSE, purl=TRUE, echo=FALSE}
## Challenge with boxplots:
##  Start with the boxplot we created:
ggplot(data = surveys_complete, aes(x = species_id, y = weight)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(alpha = 0.3, color = "tomato")

##  1. Replace the box plot with a violin plot; see `geom_violin()`.

##  2. Represent weight on the log10 scale; see `scale_y_log10()`.

##  3. Create boxplot for `hindfoot_length` overlaid on a jitter layer.
```

## Customizing your plot

### **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white, **`ggplot2`**
comes with several other themes which can be useful to quickly change the look
of your visualization. The complete list of themes is available
at <http://docs.ggplot2.org/current/ggtheme.html>. `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)
package provides a wide variety of options (including an Excel 2003 theme). The
[**`ggplot2`** extensions website](https://www.ggplot2-exts.org) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.


### Changing the axes

* xlab
* ylab
* ggtitle

### Changing the legend

* labels
* legend position
* no need to go into too much details here, refer to cookbook.

## Arranging and Exporting 

* Talk about patchwork? not on CRAN yet though
* Show how to use ggsave






```{r}
library(ratdat)


portal <- dplyr::left_join(surveys, species) %>%
    dplyr::left_join(plots) %>%
    dplyr::filter(sex %in% c("M", "F")) %>%
    dplyr::filter(!is.na(hindfoot_length),
                  !is.na(weight)) %>% 
    dplyr::filter(taxa == "Rodent")


portal_dipo <- portal %>%
    dplyr::filter(species_id %in% c("DO", "DM", "DS"),
                  year >= 1977 & year <= 1982)
```


```{r, child="_page_built_on.Rmd"}
```
