---
layout: topic
title: Working With Vectors
author: Steve Pederson
minutes: 10
---

Before moving onto the `survey` data we saw in yesterday's SQL session, let's spend 10 more minutes on vectors.
We'll look at:

- Vector types
- Coercion
- Subsetting vectors

We know the four most common atomic vector types:

1. logical  
2. integer  
3. double  
4. character  

Let's create a few, using the `c()` command we've just seen:

---

# Vector Types

## A Logical Vector
```{r, eval=TRUE,  purl=FALSE}
logi_vec  <- c(TRUE, TRUE, FALSE)
class(logi_vec)
typeof(logi_vec)
```

---

## An Integer Vector
```{r, eval=TRUE,  purl=FALSE}
int_vec <- 1:5
class(int_vec)
typeof(int_vec)
```

---

## A Double Vector

These are often **lazily** referred to as numeric vectors.
Both `double` and `integer` vectors are numeric.

```{r, eval=TRUE,  purl=FALSE}
dbl_vec <- c(0.618, 1.414, 2)
class(dbl_vec)
typeof(dbl_vec)
```

This is why `typeof()` is preferable for checking vectors.

---

## A Character Vector
```{r, eval=TRUE,  purl=FALSE}
char_vec <- c("a", "cat", "video")
class(char_vec)
typeof(char_vec)
```

---
# Coersion

Can we convert these vectors into another type of value?
There are a series of functions `as.logical()`, `as.integer()`, `as.double()` and `as.character()`

```{r}
as.integer(logi_vec)
as.logical(int_vec)
as.character(logi_vec)
```

---

# Questions
What do you think will happen if you try?
```{r, eval=FALSE}
as.integer(char_vec)
```

What type of vector is this?
```{r, echo=FALSE, include=FALSE}
sim_vec <- c(742, "Evergreen", "Tce")
```


```{r, eval=FALSE}
sim_vec <- c(742, "Evergreen", "Tce")
```

What would happen if we tried to convert this to a numeric vector?
```{r, eval=FALSE}
as.numeric(sim_vec)
```

## Combining Vectors of different types
```{r}
new_vec <- c(dbl_vec, int_vec)
typeof(new_vec)
```

### More questions
What would happen if combining the following vectors?
```{r, eval=FALSE}
logi_int_vec  <- c(logi_vec, int_vec)
typeof(what_vec)
int_char_vec  <- c(int_vec, char_vec)
typeof(int_char_vec)
```


# Subsetting vectors
In R we use square brackets to extract single values, or a **subset** of the values
```{r}
char_vec[2]
```

We can use the `minus` sign to extract all values, *except* a given value
```{r}
char_vec[-2]
```

We can also use a vector to do this
```{r}
char_vec[c(1, 3)]
```

And we can also rearrange a vector
```{r}
char_vec[c(3, 1, 2)]
```

## Subsetting a vector using logical values
```{r}
sim_vec[logi_vec]
```

## A common error 
What is wrong with this?
```{r, eval=FALSE}
sim_vec[int_vec]
```
