---
title: Starting with data
author: Data Carpentry contributors
minutes: 20
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ## Learning Objectives
>
> * Describe what a data.frame is.
> * Summarize the contents of a dataframe in R.
> * Load external data from a .csv file into R.
> * Manipulate categorical data in R.

------------

## Presentation of the Survey Data

```{r, echo=FALSE, purl=TRUE}
### Presentation of the survey data
```

We are studying the species and weight of animals caught in plots in our study
area. The dataset is stored as a comma separated value (CSV) file.
Each row holds information for a single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxa             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from figshare, and we will use `read.csv()` to
load into memory the content of the CSV file as a `data.frame`.

To download the data into the `data/` subdirectory, do:

```{r, eval=FALSE, purl=TRUE}
download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data_joined.csv")
```

You are now ready to load the data:

```{r, eval=TRUE,  purl=FALSE}
surveys <- read.csv('data/portal_data_joined.csv')
```

This statement doesn't produce any output because, as you might recall,
assignment doesn't display anything. If we want to check that our data has been
loaded, we can print the variable's value: `surveys`.

Wow... that was a lot of output. At least it means the data loaded
properly. Let's check the top (the first 6 lines) of this `data.frame` using the
function `head()`:

```{r, results='show', purl=FALSE}
head(surveys)
```

A `data.frame` is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because the column are
vectors, they all contain the same type of data. We can see this when inspecting
the __str__ucture of a `data.frame` with the function `str()`:

```{r, purl=FALSE}
str(surveys)
```

### Challenge

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?

```{r, echo=FALSE, purl=TRUE}

## Challenge
## Based on the output of `str(surveys)`, can you answer the following questions?
## * What is the class of the object `surveys`?
## * How many rows and how many columns are in this object?
## * How many species have been recorded during these surveys?

```

<!---
```{r, echo=FALSE, purl=FALSE}
## Answers
##
## * class: data frame
## * how many rows: 34786,  how many columns: 13
## * how many species: 48
```
--->

## Inspecting `data.frame` Objects

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a `data.frame`. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data.

* Size:
    * `dim(surveys)` - returns a vector with the number of rows in the first element,
          and the number of columns as the second element (the **dim**ensions of
          the object)
    * `nrow(surveys)` - returns the number of rows
    * `ncol(surveys)` - returns the number of columns

* Content:
    * `head(surveys)` - shows the first 6 rows
    * `tail(surveys)` - shows the last 6 rows

* Names:
    * `names(surveys)` - returns the column names (synonym of `colnames()` for `data.frame`
	   objects)
    * `rownames(surveys)` - returns the row names

* Summary:
    * `str(surveys)` - structure of the object and information about the class, length and
	   content of  each column
    * `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.

## Indexing and subsetting data frames

```{r, echo=FALSE, purl=TRUE}

## Sequences and Subsetting data frames
```


Our survey data frame has rows and columns (it has 2 dimensions), if we want to
extract some specific data from it, we need to specify the "coordinates" we
want from it. Row numbers come first, followed by column numbers. However, note
that different ways of specifying these coordinates lead to results with
different classes.

```{r, purl=FALSE}
surveys[1]      # first column in the data frame (as a data.frame)
surveys[, 1]     # first column in the data frame (as a vector)
surveys[1, 1]   # first element in the first column of the data frame (as a vector)
surveys[1, 6]   # first element in the 6th column (as a vector)
surveys[1:3, 7] # first three elements in the 7th column (as a vector)
surveys[3, ]    # the 3rd element for all columns (as a data.frame)
head_surveys <- surveys[1:6, ] # equivalent to head(surveys)
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain parts of a data frame using the "`-`" sign:

```{r, purl=FALSE}
surveys[,-1]   #The whole data frame, except the first column
surveys[-c(7:34786),] #equivalent to head(surveys)
```

As well as using numeric values to subset a `data.frame` (or `matrix`), columns
can be called by name, using one of the three following notations:

```{r, eval = FALSE, purl=FALSE}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a vector
surveys[["species_id"]]     # Result is a vector
surveys$species_id          # Result is a vector
```

For our purposes, the last three notations are equivalent. However, the last one
with the `$` does partial matching on the name. So you could also select the
column `"day"` by typing `surveys$d`. It's a shortcut, as with all shortcuts,
they can have dangerous consequences, and are best avoided. Besides, with
auto-completion in RStudio, you rarely have to type more than a few characters
to get the full and correct column name.

> ### Challenge
>
> 1. Create a `data.frame` (`surveys_200`) containing only the observations from row 200 of the
>    `surveys` dataset.
>
> 2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
>
>      * Use that number to pull out just that last row in the data frame
>      * Compare that with what you see as the last row using `tail()` to make
>        sure it's meeting expectations.
>      * Pull out that last row using `nrow()` instead of row number
>      * Create a new data frame object (`surveys_last`) from that last row
>      * Which
>
> 3. Use `nrow()` in conjunction with the `:` function to create a new
>    data frame called `surveys_last_100` that includes the 100 rows of the
>    `surveys` data frame.
>
> 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
>    `head(surveys)` keeping just the first through 7th rows of the surveys
>    dataset.


```{r, echo=FALSE, purl=TRUE}
### Challenges:
### 1. Create a `data.frame` (`surveys_200`) containing only the observations from row 200 of the
###    `surveys` dataset.
###
### 2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
###
###      * Use that number to pull out just that last row in the data frame
###      * Compare that with what you see as the last row using `tail()` to make
###        sure it's meeting expectations.
###      * Pull out that last row using `nrow()` instead of row number
###      * Create a new data frame object (`surveys_last`) from that last row
###      * Which
###
### 3. Use `nrow()` in conjunction with the `:` function to create a new
###    data frame called `surveys_last_100` that includes the 100 rows of the
###    `surveys` data frame.
###
### 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
###    `head(surveys)` keeping just the first through 7th rows of the surveys
###    dataset.

```

<!---
```{r, purl=FALSE}
## Answers
surveys_1999 <- surveys[200, ]
surveys_last <- surveys[nrow(surveys), ]
surveys_last_100 <- surveys[(nrow(surveys)-99):nrow(surveys), ]
surveys_head <- surveys[-c(7:nrow(surveys)),]
```
--->


## Factors

```{r, echo=FALSE, purl=TRUE}
### Factors
```

When we did `str(surveys)` we see that several of the columns consist of
integers, however, the columns `genus`, `species`, `sex`, `plot_type`, ... are
of a special class called a `factor`. Factor are very useful and are actually
something that make R particularly well suited to working with data, so we're
going to spend a little time introducing them.

Factors are used to represent categorical data. Factors can be ordered or
unordered, and understanding them is necessary for statistical analysis and for
plotting.

Factors are stored as integers, and have labels associated with these unique
integers. While factors look (and often behave) like character vectors, they are
actually integers under the hood, and you need to be careful when treating them
like strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts *levels* in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because
`f` comes before `m`, even though the first element in this vector is
`"male"`). You can check this by using the function `levels()`, and check the
number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `sex` vector would be:

```{r, results=TRUE, purl=FALSE}
sex
sex <- factor(sex, levels = c("male", "female"))
sex
```

In R's memory, these factors are represented by integers (1, 2, 3), but are more
informative than integers because factors are self describing: `"female"`,
`"male"` is more descriptive than `1`, `2`. Which one is "male"?  You wouldn't
be able to tell just from the integer data. Factors, on the other hand, have
this information built in. It is particularly helpful when there are many levels
(like the species names in our example data set).

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(sex)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier.  One method is to
convert factors to characters and then numbers.  Another method is to use the
`levels()` function. Compare:

```{r, purl=TRUE}
f <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
as.numeric(levels(f))[f]    ## The recommended way.
```

Notice that in the `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(f)`
* We convert these levels to numeric values using `as.numeric(levels(f))`
* We then access these numeric values using the underlying integers of the
  vector `f` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor
level. Let's look at the number of males and females captured over the course of
the experiment:

```{r, purl=TRUE}
## bar plot of the number of females and males captured during the experiment:
plot(surveys$sex)
```

In addition of males and females, there are about 1700 individuals for which the
sex information hasn't been recorded. Additionally, for the these individuals,
there is no label to indicate that the information is missing. Let's rename this
label to something more meaningful. Before doing that, we're going to pull out
the data on sex and work with that data, so we're not modifying the working copy
of the data frame:

```{r, purl=FALSE}
sex <- surveys$sex
head(sex)
levels(sex)
levels(sex)[1] <- "missing"
levels(sex)
head(sex)
```


### Challenges

* Rename "F" and "M" to "female" and "male" respectively.
* Now that we have renamed the factor level to "missing", can you recreate the
  barplot such that "missing" is last (after "male")


```{r wrong-order, results='show', echo=FALSE, purl=TRUE}
## Challenges
##
## * Rename "F" and "M" to "female" and "male" respectively.
## * Now that we have renamed the factor level to "missing", can you recreate the
##   barplot such that "missing" is last (after "male")
```


<!---
```{r correct-order, purl=FALSE}
## Answers
levels(sex)[2:3] <- c("female", "male")
sex <- factor(sex, levels = c("female", "male", "missing"))
plot(sex)
```
--->
