---
title: "Starting with data"
author: "Data Carpentry contributors"
minutes: 20
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ### Learning Objectives
>
> * Describe what a data frame is.
> * Load external data from a .csv file into a data frame in R.
> * Summarize the contents of a data frame in R.
> * Manipulate categorical data in R.
> * Change how character strings are handled in a data frame.
> * Format dates in R

------------

## Presentation of the Survey Data

```{r, echo=FALSE, purl=TRUE}
### Presentation of the survey data
```

We are studying the species and weight of animals caught in plots in our study
area. The dataset is stored as a comma separated value (CSV) file.
Each row holds information for a single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxa             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from figshare, and we will use `read.csv()` to
load into memory the content of the CSV file as an object of class `data.frame`.

To download the data into the `data/` subdirectory, run the following:

```{r, eval=FALSE, purl=TRUE}
download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data_joined.csv")
```

You are now ready to load the data:

```{r, eval=TRUE,  purl=FALSE}
surveys <- read.csv('data/portal_data_joined.csv')
```

This statement doesn't produce any output because, as you might recall,
assignments don't display anything. If we want to check that our data has been
loaded, we can print the variable's value: `surveys`.

Wow... that was a lot of output. At least it means the data loaded
properly. Let's check the top (the first 6 lines) of this data frame using the
function `head()`:

```{r, results='show', purl=FALSE}
head(surveys)
```

## What are data frames?

Data frames are the _de facto_ data structure for most tabular data, and what we
use for statistics and plotting.

A data frame can be created by hand, but most commonly they are generated by the
functions `read.csv()` or `read.table()`; in other words, when importing
spreadsheets from your hard drive (or the web).

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because the column are
vectors, they all contain the same type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame comprising of a
numeric, a character and a logical vector.

![](./img/data-frame.svg)


We can see this when inspecting the <b>str</b>ucture of a data frame
with the function `str()`:

```{r, purl=FALSE}
str(surveys)
```

## Inspecting `data.frame` Objects

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a data frame. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data. Let's try them out!

* Size:
    * `dim(surveys)` - returns a vector with the number of rows in the first element,
          and the number of columns as the second element (the **dim**ensions of
          the object)
    * `nrow(surveys)` - returns the number of rows
    * `ncol(surveys)` - returns the number of columns

* Content:
    * `head(surveys)` - shows the first 6 rows
    * `tail(surveys)` - shows the last 6 rows

* Names:
    * `names(surveys)` - returns the column names (synonym of `colnames()` for `data.frame`
	   objects)
    * `rownames(surveys)` - returns the row names

* Summary:
    * `str(surveys)` - structure of the object and information about the class, length and
	   content of  each column
    * `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.


> ### Challenge
>
> Based on the output of `str(surveys)`, can you answer the following questions?
>
> * What is the class of the object `surveys`?
> * How many rows and how many columns are in this object?
> * How many species have been recorded during these surveys?

```{r, echo=FALSE, purl=TRUE}

## Challenge
## Based on the output of `str(surveys)`, can you answer the following questions?
## * What is the class of the object `surveys`?
## * How many rows and how many columns are in this object?
## * How many species have been recorded during these surveys?

```

<!---
```{r, echo=FALSE, purl=FALSE}
## Answers
##
## * class: data frame
## * how many rows: 34786,  how many columns: 13
## * how many species: 48
```
--->


## Indexing and subsetting data frames

```{r, echo=FALSE, purl=TRUE}

## Indexing and subsetting data frames
```


Our survey data frame has rows and columns (it has 2 dimensions), if we want to
extract some specific data from it, we need to specify the "coordinates" we
want from it. Row numbers come first, followed by column numbers. However, note
that different ways of specifying these coordinates lead to results with
different classes.

```{r, purl=FALSE}
surveys[1, 1]   # first element in the first column of the data frame (as a vector)
surveys[1, 6]   # first element in the 6th column (as a vector)
surveys[, 1]    # first column in the data frame (as a vector)
surveys[1]      # first column in the data frame (as a data.frame)
surveys[1:3, 7] # first three elements in the 7th column (as a vector)
surveys[3, ]    # the 3rd element for all columns (as a data.frame)
head_surveys <- surveys[1:6, ] # equivalent to head(surveys)
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain parts of a data frame using the "`-`" sign:

```{r, purl=FALSE}
surveys[,-1]          # The whole data frame, except the first column
surveys[-c(7:34786),] # Equivalent to head(surveys)
```

As well as using numeric values to subset a `data.frame` (or `matrix`), columns
can be called by name, using one of the four following notations:

```{r, eval = FALSE, purl=FALSE}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a vector
surveys[["species_id"]]     # Result is a vector
surveys$species_id          # Result is a vector
```

For our purposes, the last three notations are equivalent. RStudio knows about
the columns in your data frame, so you can take advantage of the autocompletion
feature to get the full and correct column name.

> ### Challenge
>
> 1. Create a `data.frame` (`surveys_200`) containing only the observations from
>    row 200 of the `surveys` dataset.
>
> 2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
>
>      * Use that number to pull out just that last row in the data frame.
>      * Compare that with what you see as the last row using `tail()` to make
>        sure it's meeting expectations.
>      * Pull out that last row using `nrow()` instead of the row number.
>      * Create a new data frame object (`surveys_last`) from that last row.
>
> 3. Use `nrow()` to extract the row that is in the middle of the data
>    frame. Store the content of this row in an object named `surveys_middle`.
>
> 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
>    `head(surveys)` keeping just the first through 6th rows of the surveys
>    dataset.


```{r, echo=FALSE, purl=TRUE}
### Challenges:
###
### 1. Create a `data.frame` (`surveys_200`) containing only the
###    observations from row 200 of the `surveys` dataset.
###
### 2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
###
###      * Use that number to pull out just that last row in the data frame
###      * Compare that with what you see as the last row using `tail()` to make
###        sure it's meeting expectations.
###      * Pull out that last row using `nrow()` instead of the row number
###      * Create a new data frame object (`surveys_last`) from that last row
###
### 3. Use `nrow()` to extract the row that is in the middle of the
###    data frame. Store the content of this row in an object named
###    `surveys_middle`.
###
### 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
###    `head(surveys)` keeping just the first through 6th rows of the surveys
###    dataset.

```

<!---
```{r, purl=FALSE}
## Answers
surveys_200 <- surveys[200, ]
surveys_last <- surveys[nrow(surveys), ]
surveys_middle <- surveys[nrow(surveys)/2, ]
surveys_head <- surveys[-c(7:nrow(surveys)),]
```
--->


## Factors

```{r, echo=FALSE, purl=TRUE}
### Factors
```

When we did `str(surveys)` we saw that several of the columns consist of
integers, however, the columns `genus`, `species`, `sex`, `plot_type`, ... are
of a special class called a `factor`. Factors are very useful and are actually
something that make R particularly well suited to working with data, so we're
going to spend a little time introducing them.

Factors are used to represent categorical data. Factors can be ordered or
unordered, and understanding them is necessary for statistical analysis and for
plotting.

Factors are stored as integers, and have labels (text) associated with these
unique integers. While factors look (and often behave) like character vectors,
they are actually integers under the hood, and you need to be careful when
treating them like strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts *levels* in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because
`f` comes before `m`, even though the first element in this vector is
`"male"`). You can check this by using the function `levels()`, and check the
number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `sex` vector would be:

```{r, results=TRUE, purl=FALSE}
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2, 3), but are more
informative than integers because factors are self describing: `"female"`,
`"male"` is more descriptive than `1`, `2`. Which one is "male"?  You wouldn't
be able to tell just from the integer data. Factors, on the other hand, have
this information built in. It is particularly helpful when there are many levels
(like the species names in our example dataset).

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(sex)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier.  One method is to
convert factors to characters and then numbers.  Another method is to use the
`levels()` function. Compare:

```{r, purl=TRUE}
f <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(f)               # wrong! and there is no warning...
as.numeric(as.character(f)) # works...
as.numeric(levels(f))[f]    # The recommended way.
```

Notice that in the `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(f)`
* We convert these levels to numeric values using `as.numeric(levels(f))`
* We then access these numeric values using the underlying integers of the
  vector `f` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor
level. Let's look at the number of males and females captured over the course of
the experiment:

```{r, purl=TRUE}
## bar plot of the number of females and males captured during the experiment:
plot(surveys$sex)
```

In addition to males and females, there are about 1700 individuals for which the
sex information hasn't been recorded. Additionally, for these individuals,
there is no label to indicate that the information is missing. Let's rename this
label to something more meaningful. Before doing that, we're going to pull out
the data on sex and work with that data, so we're not modifying the working copy
of the data frame:

```{r, results=TRUE, purl=FALSE}
sex <- surveys$sex
head(sex)
levels(sex)
levels(sex)[1] <- "missing"
levels(sex)
head(sex)
```


> ### Challenge
>
> * Rename "F" and "M" to "female" and "male" respectively.
> * Now that we have renamed the factor level to "missing", can you recreate the
>  barplot such that "missing" is last (after "male")?


```{r wrong-order, results='show', echo=FALSE, purl=TRUE}
## Challenges
##
## * Rename "F" and "M" to "female" and "male" respectively.
## * Now that we have renamed the factor level to "missing", can you recreate the
##   barplot such that "missing" is last (after "male")
```


<!---
```{r correct-order, purl=FALSE}
## Answers
levels(sex)[2:3] <- c("female", "male")
sex <- factor(sex, levels = c("female", "male", "missing"))
plot(sex)
```
--->

### Using `stringsAsFactors=FALSE`

By default, when building or importing a data frame, the columns that contain
characters (i.e., text) are coerced (=converted) into the `factor` data
type. Depending on what you want to do with the data, you may want to keep these
columns as `character`. To do so, `read.csv()` and `read.table()` have an
argument called `stringsAsFactors` which can be set to `FALSE`.

In most cases, it's preferable to set `stringsAsFactors = FALSE` when importing
your data, and converting as a factor only the columns that require this data
type.

Compare the output of `str(surveys)` when setting `stringsAsFactors = TRUE`
(default) and `stringsAsFactors = FALSE`:


```{r, eval=FALSE, purl=FALSE}
## Compare the difference between when the data are being read as
## `factor`, and when they are being read as `character`.
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)
str(surveys)
## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)
```


> ### Challenge
>
> 1. We have seen how data frames are created when using the `read.csv()`, but
>   they can also be created by hand with the `data.frame()` function.  There are
>   a few mistakes in this hand-crafted `data.frame`, can you spot and fix them?
>   Don't hesitate to experiment!
>
>     ```{r, eval=FALSE, purl=FALSE}
>     animal_data <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
>                               feel=c("furry", "squishy", "spiny"),
>                               weight=c(45, 8 1.1, 0.8))
>     ```
>
>     ```{r, eval=FALSE, purl=TRUE, echo=FALSE}
>     ## Challenge:
>     ##  There are a few mistakes in this hand-crafted `data.frame`,
>     ##  can you spot and fix them? Don't hesitate to experiment!
>     animal_data <- data.frame(animal=c(dog, cat, sea cucumber, sea urchin),
>                               feel=c("furry", "squishy", "spiny"),
>                               weight=c(45, 8 1.1, 0.8))
>     ```
>
> 2. Can you predict the class for each of the columns in the following example?
>    Check your guesses using `str(country_climate)`:
>      * Are they what you expected?  Why? Why not?
>      * What would have been different if we had added `stringsAsFactors = FALSE` to this call?
>      * What would you need to change to ensure that each column had the accurate data type?
>
>     ```{r, eval=FALSE, purl=FALSE}
>     country_climate <- data.frame(
>            country=c("Canada", "Panama", "South Africa", "Australia"),
>            climate=c("cold", "hot", "temperate", "hot/temperate"),
>            temperature=c(10, 30, 18, "15"),
>            northern_hemisphere=c(TRUE, TRUE, FALSE, "FALSE"),
>            has_kangaroo=c(FALSE, FALSE, FALSE, 1)
>            )
>     ```
>
>    ```{r, eval=FALSE, purl=TRUE, echo=FALSE}
>    ## Challenge:
>    ##   Can you predict the class for each of the columns in the following
>    ##   example?
>    ##   Check your guesses using `str(country_climate)`:
>    ##   * Are they what you expected? Why? why not?
>    ##   * What would have been different if we had added `stringsAsFactors = FALSE`
>    ##     to this call?
>    ##   * What would you need to change to ensure that each column had the
>    ##     accurate data type?
>    country_climate <- data.frame(country=c("Canada", "Panama", "South Africa", "Australia"),
>                                   climate=c("cold", "hot", "temperate", "hot/temperate"),
>                                   temperature=c(10, 30, 18, "15"),
>                                   northern_hemisphere=c(TRUE, TRUE, FALSE, "FALSE"),
>                                   has_kangaroo=c(FALSE, FALSE, FALSE, 1))
>    ```
>
>    <!--- Answers
>
>    ```{r, eval=FALSE, echo=FALSE, purl=FALSE}
>    ## Answers
>    ## * missing quotations around the names of the animals
>    ## * missing one entry in the "feel" column (probably for one of the furry animals)
>    ## * missing one comma in the weight column
>
>    ## Answers
>    ## * `country`, `climate`, `temperature`, and `northern_hemisphere` are
>    ##    factors; `has_kangaroo` is numeric.
>    ## * using `stringsAsFactors=FALSE` would have made them character instead of
>    ##   factors
>    ## * removing the quotes in temperature, northern_hemisphere, and replacing 1
>    ##   by TRUE in the `has_kangaroo` column would probably what was originally
>    ##   intended.
>    ```
>
>    -->
>

The automatic conversion of data type is sometimes a blessing, sometimes an
annoyance. Be aware that it exists, learn the rules, and double check that data
you import in R are of the correct type within your data frame. If not, use it
to your advantage to detect mistakes that might have been introduced during data
entry (a letter in a column that should only contain numbers for instance).

## Formatting Dates

One of the most common issues that new (and experienced!) R users have is converting
date and time information into a variable that is appropriate and usable during
analyses. As a reminder from earlier in this lesson, the best practice for dealing
with date data is to ensure that each component of your date is stored as a separate
variable. Using `str()`, We can confirm that our data frame has a separate column for
day, month, and year, and each contains integer values.

```{r, eval=FALSE, purl=FALSE}
str(surveys)
```

We're going to be using the `ymd()` function from the package **`lubridate`**. This
function is designed to take a vector representing year, month, and day and convert
that information to a POSIXct vector. POSIXct is a class of data recognized by R as
being a date or date and time. The argument that the function requires is relatively
flexible, but, as a best practice, is a character vector formatted as "YYYY-MM-DD".

Start by loading the required package:

```{r load-package, message=FALSE, purl=FALSE}
library(lubridate)
```

Create a character vector from the `year`, `month`, and `day` columns of `surveys` using `paste()`:

```{r, eval=FALSE, purl=FALSE}
paste(surveys$year, surveys$month, surveys$day, sep="-")
# sep indicates the character to use to separate each component
```

This character vector can be used as the argument for `ymd()`:

```{r, eval=FALSE, purl=FALSE}
ymd(paste(surveys$year, surveys$month, surveys$day, sep="-"))
```

The resulting POSIXct vector can be added to `surveys` as a new column called `date`:

```{r, eval=FALSE, purl=FALSE}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep="-"))

str(surveys) # notice the new column, with 'date' as the class
```

<p style="text-align: right; font-size: small;">Page build on: `r format(Sys.time())`</p>
