---
layout: topic
title: Manipulating data
minutes: 45
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
surveys <- read.csv(file="../../data/biology/surveys.csv")
```

> ## Learning Objectives
>
> * being able to subset for a level
> * being able to extract particular values from a `data.frame`
> * being able to add/remove columns and rows


# Subsetting data

```{r, echo=FALSE, purl=TRUE}
## subsetting data
```

In particular for larger datasets, it can be tricky to remember the column
number that corresponds to a particular variable. (Are species names in column 5
or 7? oh, right... they are in column 6). In some cases, in which column the
variable will be can change if the script you are using adds or removes
columns. It's therefore often better to use column names to refer to a
particular variable, and it makes your code easier to read and your intentions
clearer.

You can do operations on a particular column, by selecting it using the `$`
sign. In this case, the entire column is a vector. For instance, to extract all
the weights from our datasets, we can use: `surveys$wgt`. You can use
`names(surveys)` or `colnames(surveys)` to remind yourself of the column names.

In some cases, you may way to select more than one column. You can do this using
the square brackets: `surveys[, c("wgt", "sex")]`.

When analyzing data, though, we often want to look at partial statistics, such
as the maximum value of a variable per species or the average value per plot.

One way to do this is to select the data we want, and create a new temporary
array, using the `subset()` function. For instance, if we just want to look at
the animals of the species "DO":

```{r, purl=FALSE}
surveys_DO <- subset(surveys, species == "DO")
```

### Challenge

1. What does the following do?

  ```{r, echo=FALSE, purl=TRUE}
  ## What does the following do? (Try to guess without executing it)
  ## surveys_DO$month[2] <- 8
  ```

1. Use the function `subset` twice to create a `data.frame` that contains all
individuals of the species "DM" that were collected in 2002.
  * How many individuals of the species "DM" were collected in 2002?

  ```{r, echo=FALSE, purl=TRUE}
  ##  Use the function `subset` twice to create a `data.frame` that
  ##  contains all individuals of the species "DM" that were collected
  ##  in 2002.
  ##   * How many individuals of the species "DM" were collected in 2002?
  ```

## Adding a column to our dataset

```{r, echo=FALSE, purl=TRUE}
## Adding columns
```

Sometimes, you may have to add a new column to your dataset that represents a
new variable. You can add columns to a `data.frame` using the function `cbind()`
(__c__olumn __bind__). Beware, the additional column must have the same number
of elements as there are rows in the `data.frame`.

In our survey dataset, the species are represented by a 2-letter code (e.g.,
"AB"), however, we would like to include the species name. The correspondance
between the 2 letter code and the names are in the file `species.csv`. In this
file, one column includes the genus and another includes the species. First, we
are going to import this file in memory:

```{r, echo=FALSE, purl=FALSE}
species <- read.csv("../../data/biology/species.csv")
```

```{r, eval=FALSE, purl=TRUE}
species <- read.csv("data/species.csv")
```

We are then going to use the function `match()` to create a vector that contains
the genus names for all our observations. The function `match()` takes at least
2 arguments: the values to be matched (in our case the 2 letter code from the
`surveys` data frame held in the column `species`), and the table that contains
the values to be matched against (in our case the 2 letter code in the `species`
data frame held in the column `species_id`). The function returns the position
of the matches in the table, and this can be used to retrieve the genus names:

```{r, purl=TRUE}
surveys_spid_index <- match(surveys$species, species$species_id)
surveys_genera <- species$genus[surveys_spid_index]
```

Now that we have our vector of genus names, we can add it as a new column to our
`surveys` object:

```{r, purl=TRUE}
surveys <- cbind(surveys, genus=surveys_genera)
```

### Challenge

Use the same approach to also include the species names in the `surveys` data
frame.

```{r, echo=FALSE, purl=TRUE}
## Use the same approach to also include the species names in the
## `surveys` data frame.
```


```{r, echo=FALSE, purl=FALSE}
surveys_species <- species$species[surveys_spid_index]
surveys <- cbind(surveys, species_name=surveys_species)
```

```{r, results='show'}
## and check out the result
head(surveys)
```

<!--- should we cover merge()? --->

# Adding rows

<!--- Even if this is not optimal, using this approach requires to cover less   -->
<!--- material such as logical operations on vectors. Depending on how fast the -->
<!--- group moves, it might be better to show the correct way.                  -->

```{r, echo=FALSE, purl=TRUE}
## Adding rows
```

Let's create a `data.frame` that contains the information only for the species
"DO" and "DM". We know how to create the data set for each species with the
function `subset()`:

```{r, purl=FALSE}
surveys_DO <- subset(surveys, species == "DO")
surveys_DM <- subset(surveys, species == "DM")
```

Similarly to `cbind()` for columns, there is a function `rbind()` (__r__ow
__bind__) that puts together two `data.frame`. With `rbind()` the number of
columns and their names must be identical between the two objects:

```{r, purl=FALSE}
surveys_DO_DM <- rbind(surveys_DO, surveys_DM)
```

### Challenge

Using a similar approach, construct a new `data.frame` that only includes data
for the years 2000 and 2001.

```{r, echo=FALSE, purl=TRUE}
## How many columns are now in (1) the `data.frame` `surveys`, (2) the `data.frame`
## `surveys_index`?
```

# Removing columns

```{r, echo=FALSE, purl=FALSE}
## Removing columns
```

Just like you can select columns by their positions in the `data.frame` or by
their names, you can remove them similarly.

To remove it by column number:

```{r, results='show', purl=FALSE}
surveys_noDate <- surveys[, -c(2:4)]
colnames(surveys)
colnames(surveys_noDate)
```

The easiest way to remove by name is to use the `subset()` function. This time
we need to specify explicitly the argument `select` as the default is to subset
on rows (as above). The minus sign indicates the names of the columns to remove
(note that the column names should not be quoted):

```{r, results='show', purl=FALSE}
surveys_noDate2 <- subset(surveys, select=-c(month, day, year))
colnames(surveys_noDate2)
```

# Removing rows

```{r, echo=FALSE, purl=FALSE}
## Removing rows
```

Typically rows are not associated with names, so to remove them from the
`data.frame`, you can do:

```{r, results='show', purl=FALSE}
surveys_missingRows <- surveys[-c(10, 50:70), ] # removing rows 10, and 50 to 70
```
