---
layout: topic
title: Analyzing and Plotting Data
minutes: 60
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
surveys <- read.csv(file="../../data/biology/surveys.csv")
```

> ## Learning Objectives
>
> * understand how to deal with missing data
> * being able to generate summary statistics from the data
> * calculate basic statistics across a levels of a factor (here species)
> * generate plot from the data using the average weights of the animals as an
>   example


# Calculating statistics

```{r, echo=FALSE, purl=FALSE}
## Calculating statistics
```

Let's get a closer look at our data. For instance, we might want to know how
many animals we trapped in each plot, or how many of each species were caught.

To get a `vector` of all the species, we are going to use the `unique()`
function that tells us the unique values in a given vector:

```{r, purl=FALSE}
unique(surveys$species)
```

The function `table()`, tells us how many of each species we have:

```{r, purl=FALSE}
table(surveys$species)
```

R has a lot of built in statistical functions, like `mean()`, `median()`,
`max()`, `min()`. Let's start by calculating the average weight of all the
animals using the function `mean()`:

```{r, results='show', purl=FALSE}
mean(surveys$wgt)
```

Hmm, we just get `NA`. That's because we don't have the weight for every animal
and missing data is recorded as `NA`. By default, all R functions operating on a
vector that contains missing data will return NA. It's a way to make sure that
users know they have missing data, and make a conscious decision on how to deal
with it.

When dealing with simple statistics like the mean, the easiest way to ignore
`NA` (the missing data) is to use `na.rm=TRUE` (`rm` stands for remove):

```{r, results='show', purl=FALSE}
mean(surveys$wgt, na.rm=TRUE)
```

In some cases, it might be useful to remove the missing data from the
vector. For this purpose, R comes with the function `na.omit`:

```{r, purl=FALSE}
wgt_noNA <- na.omit(surveys$wgt)
```

For some applications, it's useful to keep all observations, for others, it
might be best to remove all observations that contain missing data. The function
`complete.cases()` removes any rows that contain at least one missing
observation:

```{r, purl=FALSE}
surveys_complete <- surveys[complete.cases(surveys), ]
```

<!--- need to cover negation, and vector operations for this...
If you want to remove only the observations that are missing data for one
variable, you can use the function `is.na()`. For instance, to create a new
dataset that only contains individuals that have been weighted:

```{r, purl=FALSE}
surveys_with_weights <- surveys[!is.na(surveys$weight), ]
```
--->

### Challenge

1. To determine the number of elements found in a vector, we can use
use the function `length()` (e.g., `length(surveys$wgt)`). Using `length()`, how
many animals have not had their weights recorded?

1. What is the median weight for the males?

1. What is the range (minimum and maximum) weight?

1. Bonus question: what is the standard error for the weight? (hints: there is
   no built-in function to compute standard errors, and the function for the
   square root is `sqrt()`).

```{r, echo=FALSE, purl=TRUE}
## 1. To determine the number of elements found in a vector, we can use
## use the function `length()` (e.g., `length(surveys$wgt)`). Using `length()`, how
## many animals have not had their weights recorded?

## 2. What is the median weight for the males?

## 3. What is the range (minimum and maximum) weight?

## 4. Bonus question: what is the standard error for the weight? (hints: there is
##    no built-in function to compute standard errors, and the function for the
##    square root is `sqrt()`).
```

# Statistics across factor levels

```{r, echo=FALSE, purl=TRUE}
## Statistics across factor levels
```

What if we want the maximum weight for all animals, or the average for each
plot?

R comes with convenient functions to do this kind of operations, functions in
the `apply` family.

For instance, `tapply()` allows us to repeat a function across each level of a
factor. The format is:

```{r, eval=FALSE, purl=FALSE}
tapply(columns_to_do_the_calculations_on, factor_to_sort_on, function)
```

If we want to calculate the mean for each species (using the complete dataset):

```{r, purl=FALSE}
tapply(surveys_complete$wgt, surveys_complete$species, mean)
```

This produces some `NA` because R "remembers" all species that were found in the
original dataset, even if they didn't have any weight data associated with them
in the current dataset. To remove the `NA` and make things clearer, we can
redefine the levels for the factor "species" before calculating the means. Let's
also create an object to store these values:

```{r, purl=FALSE}
surveys_complete$species <- factor(surveys_complete$species)
species_mean <- tapply(surveys_complete$wgt, surveys_complete$species, mean)
```

### Challenge

1. Create new objects to store: the standard deviation, the maximum and minimum
   values for the weight of each species
1. How many species do you have these statistics for?
1. Create a new data frame (called `surveys_summary`) that contains as columns:
   * `species` the 2 letter code for the species names
   * `mean_wgt` the mean weight for each species
   * `sd_wgt` the standard deviation for each species
   * `min_wgt`  the minimum weight for each species
   * `max_wgt`  the maximum weight for each species

```{r, echo=FALSE, purl=TRUE}
## 1. Create new objects to store: the standard deviation, the maximum and minimum
##    values for the weight of each species
## 2. How many species do you have these statistics for?
## 3. Create a new data frame (called `surveys_summary`) that contains as columns:
##    * `species` the 2 letter code for the species names
##    * `mean_wgt` the mean weight for each species
##    * `sd_wgt` the standard deviation for each species
##    * `min_wgt`  the minimum weight for each species
##    * `max_wgt`  the maximum weight for each species
```

**Answers**

```{r, purl=FALSE}
species_max <- tapply(surveys_complete$wgt, surveys_complete$species, max)
species_min <- tapply(surveys_complete$wgt, surveys_complete$species, min)
species_sd <- tapply(surveys_complete$wgt, surveys_complete$species, sd)
nlevels(surveys_complete$species) # or length(species_mean)
surveys_summary <- data.frame(species=levels(surveys_complete$species),
                              mean_wgt=species_mean,
                              sd_wgt=species_sd,
                              min_wgt=species_min,
                              max_wgt=species_max)
```

```{r, echo=FALSE, purl=TRUE}
## Plotting
```

# Creating a barplot

The mathematician Richard Hamming once said, "The purpose of computing is
insight, not numbers", and the best way to develop insight is often to visualize
data. Visualization deserves an entire lecture (or course) of its own, but we
can explore a few features of R's base plotting package.

Let's use the `surveys_summary` data that we generated and plot it.

R has built in plotting functions.

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt)
```

The axis labels are too big though, so you can't see them all. Let's change that.

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt, cex.names=0.4)
```

Alternatively, we may want to flip the axes to have more room for the species names:

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt, horiz=TRUE, las=1)
```

Let's also add some colors, and add a main title, label the axis:

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt, horiz=TRUE, las=1,
        col=c("lavender", "lightblue"), xlab="Weight (g)",
        main="Mean weight per species")
```

### Challenge

1. Create a new plot showing the standard deviation for each species. Choose
   one or more colors from
   [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf). (If you
   prefer, you can also specify colors using their hexadecimal values
   `#RRGGBB`.)

```{r, echo=FALSE, purl=TRUE}
## 1. Create a new plot showing the standard deviation for each species.
```

## More about plotting

There are lots of different ways to plot things. You can do `plot(object)` for
most classes included in R base. To explore some of the possibilities:

```{r, eval=FALSE, purl=FALSE}
?barplot
?boxplot
?plot.default
example(barplot)
```

There's also a plotting package called [`ggplot2`](http://docs.ggplot2.org/)
that adds a lot of functionality. The syntax takes some getting used to but it's
extremely powerful and flexible.

If you wanted to output this plot to a pdf file rather than to the screen, you
can specify where you want the plot to go with the `pdf()` function. If you
wanted it to be a JPG, you would use the function `jpeg()` (other formats
available: svg, png, ps).

Be sure to add `dev.off()` at the end to finalize the file. For `pdf()`, you can
create multiple pages to your file, by generating multiple plots before calling
`dev.off()`.


```{r, eval=FALSE, purl=TRUE}
pdf("mean_per_species.pdf")
barplot(surveys_summary$mean_wgt, horiz=TRUE, las=1,
        col=c("lavender", "lightblue"), xlab="Weight (g)",
        main="Mean weight per species")
dev.off()
```
