```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

# Data Carpentry R materials - Analyzing data

--------------------------------------------------

# Objectives

* understand how to deal with missing data
* being able to generate summary statistics from the data
* calculate basic statistics across a levels of a factor (here species)

--------------------------------------------------

# Calculating statistics

```{r, echo=FALSE, purl=FALSE}
## Calculating statistics
```

Let's get a closer look at our data. For instance, we might want to know how
many animals we trapped in each plot, or how many of each species were caught.

To get a `vector` of all the species, we are going to use the `unique()`
function that tells us the unique values in a given vector:

```{r, purl=FALSE}
unique(surveys$species)
```

The function `table()`, tells us how many of each species we have:

```{r, purl=FALSE}
table(surveys$species)
```

R has a lot of built in statistical functions, like `mean()`, `median()`,
`max()`, `min()`. Let's start by calculating the average weight of all the
animals using the function `mean()`:

```{r, results='show', purl=FALSE}
mean(surveys$wgt)
```

Hmm, we just get `NA`. That's because we don't have the weight for every animal
and missing data is recorded as `NA`. By default, all R functions operating on a
vector that contains missing data will return NA. It's a way to make sure that
users know they have missing data, and make a conscious decision on how to deal
with it.

When dealing with simple statistics like the mean, the easiest way to ignore
`NA` (the missing data) is to use `na.rm=TRUE` (`rm` stands for remove):

```{r, results='show', purl=FALSE}
mean(surveys$wgt, na.rm=TRUE)
```

In some cases, it might be useful to remove the missing data from the
vector. For this purpose, R comes with the function `na.omit`:

```{r, purl=FALSE}
wgt_noNA <- na.omit(surveys$wgt)
```

For some applications, it's useful to keep all observations, for others, it
might be best to remove all observations that contain missing data. The function
`complete.cases()` removes any rows that contain at least one missing
observation:

```{r, purl=FALSE}
surveys_complete <- surveys[complete.cases(surveys), ]
```

<!--- need to cover negation, and vector operations for this...
If you want to remove only the observations that are missing data for one
variable, you can use the function `is.na()`. For instance, to create a new
dataset that only contains individuals that have been weighted:

```{r, purl=FALSE}
surveys_with_weights <- surveys[!is.na(surveys$weight), ]
```
--->

### Challenge

1. To determine the number of elements found in a vector, we can use
use the function `length()` (e.g., `length(surveys$wgt)`). Using `length()`, how
many animals have not had their weights recorded?

1. What is the median weight for the males?

1. What is the range (minimum and maximum) weight?

1. Bonus question: what is the standard error for the weight? (hints: there is
   no built-in function to compute standard errors, and the function for the
   square root is `sqrt()`).

```{r, echo=FALSE, purl=TRUE}
## 1. To determine the number of elements found in a vector, we can use
## use the function `length()` (e.g., `length(surveys$wgt)`). Using `length()`, how
## many animals have not had their weights recorded?

## 2. What is the median weight for the males?

## 3. What is the range (minimum and maximum) weight?

## 4. Bonus question: what is the standard error for the weight? (hints: there is
##    no built-in function to compute standard errors, and the function for the
##    square root is `sqrt()`).
```

# Statistics across factor levels

```{r, echo=FALSE, purl=TRUE}
## Statistics across factor levels
```

What if we want the maximum weight for all animals, or the average for each
plot?

R comes with convenient functions to do this kind of operations, functions in
the `apply` family.

For instance, `tapply()` allows us to repeat a function across each level of a
factor. The format is:

```{r, eval=FALSE, purl=FALSE}
tapply(columns_to_do_the_calculations_on, factor_to_sort_on, function)
```

If we want to calculate the mean for each species (using the complete dataset):

```{r, purl=FALSE}
tapply(surveys_complete$wgt, surveys_complete$species, mean)
```

This produces some `NA` because R "remembers" all species that were found in the
original dataset, even if they didn't have any weight data associated with them
in the current dataset. To remove the `NA` and make things clearer, we can
redefine the levels for the factor "species" before calculating the means. Let's
also create an object to store these values:

```{r, purl=FALSE}
surveys_complete$species <- factor(surveys_complete$species)
species_mean <- tapply(surveys_complete$wgt, surveys_complete$species, mean)
```

### Challenge

1. Create new objects to store: the standard deviation, the maximum and minimum
   values for the weight of each species
1. How many species do you have these statistics for?
1. Create a new data frame (called `surveys_summary`) that contains as columns:
   * `species` the 2 letter code for the species names
   * `mean_wgt` the mean weight for each species
   * `sd_wgt` the standard deviation for each species
   * `min_wgt`  the minimum weight for each species
   * `max_wgt`  the maximum weight for each species

```{r, echo=FALSE, purl=TRUE}
## 1. Create new objects to store: the standard deviation, the maximum and minimum
##    values for the weight of each species
## 2. How many species do you have these statistics for?
## 3. Create a new data frame (called `surveys_summary`) that contains as columns:
##    * `species` the 2 letter code for the species names
##    * `mean_wgt` the mean weight for each species
##    * `sd_wgt` the standard deviation for each species
##    * `min_wgt`  the minimum weight for each species
##    * `max_wgt`  the maximum weight for each species
```

**Answers**

```{r, purl=FALSE}
species_max <- tapply(surveys_complete$wgt, surveys_complete$species, max)
species_min <- tapply(surveys_complete$wgt, surveys_complete$species, min)
species_sd <- tapply(surveys_complete$wgt, surveys_complete$species, sd)
nlevels(surveys_complete$species) # or length(species_mean)
surveys_summary <- data.frame(species=levels(surveys_complete$species),
                              mean_wgt=species_mean,
                              sd_wgt=species_sd,
                              min_wgt=species_min,
                              max_wgt=species_max)
```
