---
layout: topic
title: Data visualisation with ggplot2
subtitle: Visualising data in R with ggplot2 package
minutes: 60
---

<!---
show hide magic

<style> div.hidecode + pre {display: none} div.hidecode {color: #337ab7}</style><script> doclick=function(e){ e.nextSibling.nextSibling.style.display = e.nextSibling.nextSibling.style.display === "block" ? "none" : "block"; }</script>
-->

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.keep='last')
```

```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
```

Authors: **Mateusz Kuzak**, **Diana Marek**, **Hedi Peterson**



#### Disclaimer

We will here using functions of ggplot2 package. There are basic ploting
capabilities in basic R, but ggplot2 adds more powerful plotting capabilities.

> ### Learning Objectives
>
> -	Visualise some of the
>[mammals data](http://figshare.com/articles/Portal_Project_Teaching_Database/1314459)
>from Figshare [surveys.csv](http://files.figshare.com/1919744/surveys.csv)
> -	Understand how to plot these data using R ggplot2 package. For more details
>on using ggplot2 see
>[official documentation](http://docs.ggplot2.org/current/).
> -	Building step by step complex plots with ggplot2 package

Load required packages

```{r}
# plotting package
library(ggplot2)
# piping / chaining
library(magrittr)
# modern dataframe manipulations
library(dplyr)
```

Load data directly from figshare.

```{r}
surveys_raw <- read.csv("http://files.figshare.com/1919744/surveys.csv")
```

`surveys.csv` data contains some measurements of the animals caught in plots.

## Data cleaning and preparing for plotting

Let's look at the summary

```{r}
summary(surveys_raw)
```

There are few things we need to clean in the dataset.

There are missing values for species_id in some records. Let's remove those.

```{r}
surveys_complete <- surveys_raw %>%
                    filter(species_id != "")
```

We saw in summary, there were NA's in weight and hindfoot_length. Let's remove
rows with missing values in weight and hindfoot_length.
In fact, let's combine this with removing empty species_id,
so we have one command and don't make lots of intermediate variable names.
This is where piping becomes really handy!

```{r}
surveys_complete <- surveys_raw %>%
                    filter(species_id != "") %>%        # remove missing species_id
                    filter(!is.na(weight)) %>%          # remove missing weight
                    filter(!is.na(hindfoot_length))     # remove missing hindfoot_length
```

There are a lot of species with low counts, let's remove the species with less than 10 counts.

```{r}
# count records per species
species_counts <- surveys_complete %>%
                  group_by(species_id) %>%
                  tally

head(species_counts)

# get names of those frequent species
frequent_species <- species_counts %>%
                    filter(n >= 10) %>%
                    select(species_id)

surveys_complete <- surveys_complete %>%
           filter(species_id %in% frequent_species$species_id)
```

Make simple scatter plot of `hindfoot_length` (in millimeters) as a function of
`weight` (in grams), using basic R plotting capabilities.

```{r}
plot(x = surveys_complete$weight, y = surveys_complete$hindfoot_length)
```

## Plotting with ggplot2

We will make the same plot using `ggplot2` package.

`ggplot2` is a plotting package that makes it simple to create complex plots
from data in a dataframe. It uses default settings, which help creating
publication quality plots with minimal amount of settings and tweaking.

With ggplot graphics are build step by step by adding new elements.

To build a ggplot we need to:

- bind the plot to a specific data frame using the `data` argument

```{r, eval=FALSE}
ggplot(data = surveys_complete)
```

- define aesthetics (`aes`), that maps variables in the data to axes on the plot
     or to plotting size, shape color, etc.,

```{r, eval=FALSE}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length))
```

- add `geoms` -- graphical representation of the data in the plot (points,
     lines, bars). To add a geom to the plot use `+` operator:

```{r}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

Notes:

- Anything you put in the `ggplot()` function can be seen by any geom layers that you add.
i.e. these are universal plot settings
- This includes the x and y axis you set up in `aes()`

## Modifying plots

- adding transparency (alpha)

```{r}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.1)
```

- adding colors

```{r}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.1, color = "blue")
```

## Boxplot

Visualising the distribution of weight within each species.

```{r}
ggplot(data = surveys_complete, aes(x = species_id,  y = weight)) +
                   geom_boxplot()
```

By adding points to boxplot, we can see particular measurements and the
abundance of measurements.

```{r}
ggplot(data = surveys_complete, aes(x = species_id, y = weight)) +
                   geom_jitter(alpha = 0.3, color = "tomato") +
                   geom_boxplot(alpha = 0)
```

Notice how the boxplot layer is on top of the jitter layer?
Play around with the order of geoms and adjust transparency to see how to build up your plot in layers.

> ### Challenge
>
> Create boxplot for `hindfoot_length`.

## Plotting time series data

Let's calculate number of counts per year for each species. To do that we need
to group data first and count records within each group.

```{r}
yearly_counts <- surveys_complete %>%
                 group_by(year, species_id) %>%
                 tally
```

Timelapse data can be visualised as a line plot with years on x axis and counts
on y axis.

```{r}
ggplot(data = yearly_counts, aes(x = year, y = n)) +
                  geom_line()
```

Unfortunately this does not work, because we plot data for all the species
together. We need to tell ggplot to split graphed data by `species_id`

```{r}
ggplot(data = yearly_counts, aes(x = year, y = n, group = species_id)) +
  geom_line()
```

We will be able to distinguish species in the plot if we add colors.

```{r}
ggplot(data = yearly_counts, aes(x = year, y = n, group = species_id, color = species_id)) +
  geom_line()
```

## Faceting

ggplot has a special technique called *faceting* that allows to split one plot
into multiple plots based on some factor. We will use it to plot one time series
for each species separately.

```{r}
ggplot(data = yearly_counts, aes(x = year, y = n, color = species_id)) +
  geom_line() + facet_wrap(~species_id)
```

Now we would like to split line in each plot by sex of each individual
measured. To do that we need to make counts in dataframe grouped by sex.

> ### Challenges:
>
> - filter the dataframe so that we only keep records with sex "F" or "M"s
>

```{r}
sex_values = c("F", "M")
surveys_complete <- surveys_complete %>%
           filter(sex %in% sex_values)
```

> - group by year, species_id, sex

```{r}
yearly_sex_counts <- surveys_complete %>%
                     group_by(year, species_id, sex) %>%
                     tally
```

> - make the faceted plot splitting further by sex (within single plot)

```{r}
ggplot(data = yearly_sex_counts, aes(x = year, y = n, color = species_id, group = sex)) +
  geom_line() + facet_wrap(~ species_id)
```

> We can improve the plot by coloring by sex instead of species (species are
> already in separate plots, so we don't need to distinguish them better)

```{r}
ggplot(data = yearly_sex_counts, aes(x = year, y = n, color = sex, group = sex)) +
  geom_line() + facet_wrap(~ species_id)
```
