---
layout: topic
title: Data visualization with ggplot2
subtitle: Visualizing data in R with the ggplot2 package
minutes: 60
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.keep='last')
```

```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
```

#### Disclaimer

We will be using the functions in the ggplot2 package. R has powerful built-in
plotting capabilities, but for this exercise, we will be using the ggplot2
package, which facilitates the creation of highly-informative plots of
structured data.

> ### Learning Objectives
>
> -	Visualize some of the [mammals data](https://dx.doi.org/10.6084/m9.figshare.1314459.v5)
> from Figshare [surveys.csv](https://ndownloader.figshare.com/files/2292172)
> -	Understand how to plot these data using R ggplot2 package. For more details
> on using ggplot2 see [official documentation](http://docs.ggplot2.org/current/).
> -	Building step by step complex plots with the ggplot2 package

Load required packages

```{r}
# plotting package
library(ggplot2)
# piping / chaining
library(magrittr)
# modern dataframe manipulations
library(dplyr)
```

Load data directly from figshare.

```{r}
surveys_raw <- read.csv("https://ndownloader.figshare.com/files/2292172")
```

`surveys.csv` data contains some measurements of the animals caught in plots.

## Data cleaning and preparing for plotting

Let's look at the summary

```{r}
summary(surveys_raw)
```

There are few things we need to clean in the dataset.

There are missing values for species_id in some records. Let's remove those.

```{r}
surveys_complete <- surveys_raw %>%
                    filter(species_id != "")
```

We saw in summary, there were NA's in weight and hindfoot_length. Let's remove
rows with missing values in weight and hindfoot_length. In fact, let's combine
this with removing empty species_id, so we have one command and don't make lots
of intermediate variable names. This is where piping becomes really handy!

```{r}
surveys_complete <- surveys_raw %>%
                    filter(species_id != "") %>%        # remove missing species_id
                    filter(!is.na(weight)) %>%          # remove missing weight
                    filter(!is.na(hindfoot_length))     # remove missing hindfoot_length
```

There are a lot of species with low counts, let's remove the species with less
than 10 counts.

```{r}
# count records per species
species_counts <- surveys_complete %>%
                  group_by(species_id) %>%
                  tally

head(species_counts)

# get names of those frequent species
frequent_species <- species_counts %>%
                    filter(n >= 10) %>%
                    select(species_id)

surveys_complete <- surveys_complete %>%
           filter(species_id %in% frequent_species$species_id)
```

Make simple scatter plot of `hindfoot_length` (in millimeters) as a function of
`weight` (in grams), using basic R plotting capabilities.

```{r base-plot}
plot(x = surveys_complete$weight, y = surveys_complete$hindfoot_length)
```

## Plotting with ggplot2

We will make the same plot using the `ggplot2` package.

`ggplot2` is a plotting package that makes it simple to create complex plots
from data in a dataframe. It uses default settings, which help creating
publication quality plots with a minimal amount of settings and tweaking.

ggplot graphics are built step by step by adding new elements.

To build a ggplot we need to:

- bind the plot to a specific data frame using the `data` argument

```{r, eval=FALSE}
ggplot(data = surveys_complete)
```

- define aesthetics (`aes`), that maps variables in the data to axes on the plot
     or to plotting size, shape color, etc.,

```{r, eval=FALSE}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length))
```

- add `geoms` -- graphical representation of the data in the plot (points,
     lines, bars). To add a geom to the plot use `+` operator:

```{r first-ggplot}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

The `+` in the `ggplot2` package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
"templates" and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE}
# Create 
surveys_plot <- ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length))

# Draw the plot
surveys_plot + geom_point()
```

Notes:

- Anything you put in the `ggplot()` function can be seen by any geom layers that you add.
i.e. these are universal plot settings
- This includes the x and y axis you set up in `aes()`
- Aesthetics (`aes()`) can also be added separately from the `ggplot()`
  function. In other words, the last plot above can also be achieved with the
  following code:

```{r first-ggplot-alt, eval=FALSE}
ggplot(data = surveys_complete) + 
    aes(x = weight, y = hindfoot_length) +
    geom_point()
```

## Modifying plots

First, following the notes from above, let's create a ggplot "template" to
prevent unnecessary redundancy in our code.

```{r create-ggplot-object}
surveys_plot <- ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length))
```

- adding transparency (alpha)

```{r adding-transparency}
surveys_plot + geom_point(alpha = 0.1)
```

- adding colors

```{r adding-colors}
surveys_plot + geom_point(alpha = 0.1, color = "blue")
```

## Boxplot

Visualising the distribution of weight within each species.

```{r boxplot}
surveys_box <- ggplot(data = surveys_complete) + 
    aes(x = species_id, y = hindfoot_length)
surveys_box + geom_boxplot()
```

By adding points to boxplot, we can see particular measurements and the
abundance of measurements.

```{r boxplot-with-points}
surveys_box + geom_jitter(alpha = 0.3, color = "tomato") + geom_boxplot(alpha = 0)
```

Notice how the boxplot layer is on top of the jitter layer? Play around with
the order of geoms and adjust transparency to see how to build up your plot in
layers.

> ### Challenges

Boxplots are useful summaries, but hide the *shape* of the distribution. For example, if
there is a bimodal distribution, this would not be observed with a boxplot.  An alternative is
a violin plot (sometimes known as a beanplot), where the shape (of the density of points) is 
drawn.

> - Replace the box plot with a violin plot; see `geom_violin()`

In many types of data, it is important to consider the *scale* of the observations.  For example, it may be
worth changing the scale of the axis to better distribute the observations in the space of the plot.
Changing the scale of the axes is done similarly to adding/modifying other components
(i.e., by incrementally adding commands).

> - Represent weight on the log10 scale; see `scale_y_log10()`

> - Create boxplot for `hindfoot_length`.


## Plotting time series data

Let's calculate number of counts per year for each species. To do that we need
to group data first and count records within each group.

```{r}
yearly_counts <- surveys_complete %>%
                 group_by(year, species_id) %>%
                 tally
yearly_counts_plot <- ggplot(data = yearly_counts, aes(x = year, y = n))
```

Timelapse data can be visualised as a line plot with years on x axis and counts
on y axis.

```{r first-time-series}
yearly_counts_plot + geom_line()
```

Unfortunately this does not work, because we plot data for all the species
together. We need to tell ggplot to split graphed data by `species_id`. Notice
that `+` can be conveniently used to modify the default aesthetic mappings
(currently set to `x = year, y = n`).

```{r time-series-by-species}
yearly_counts_plot + aes(group = species_id) + geom_line()
```

We will be able to distinguish species in the plot if we add colors. 

```{r time-series-with-colors}
yearly_counts_plot + aes(color = species_id) + geom_line()
```

## Faceting

ggplot has a special technique called *faceting* that allows to split one plot
into multiple plots based on some factor. We will use it to plot one time
series for each species separately.

```{r first-facet}
yearly_counts_color <- yearly_counts_plot + 
    aes(group = species_id, color = species_id)
yearly_counts_color + geom_line() + facet_wrap(~species_id)
```

Now we would like to split line in each plot by sex of each individual
measured. To do that we need to make counts in dataframe grouped by sex.

> ### Challenges:
>
> - filter the dataframe so that we only keep records with sex "F" or "M"s
>

```{r}
sex_values = c("F", "M")
surveys_complete <- surveys_complete %>%
           filter(sex %in% sex_values)
```

> - group by year, species_id, sex

```{r}
yearly_sex_counts <- surveys_complete %>%
                     group_by(year, species_id, sex) %>%
                     tally
```

> - make the faceted plot splitting further by sex (within single plot)

```{r facet-by-species-and-sex}
yearly_sex_gender_plot <- ggplot(data = yearly_sex_counts) + 
    aes(x = year, y = n, color = species_id, group = sex) +
    geom_line() + 
    facet_wrap(~ species_id)

# Draw plot
yearly_sex_gender_plot
```

Usually plots with white background looks more readable in the publications.
We can set the background to white using the function `theme_bw()`.

```{r facet-by-species-and-sex-white-bg}
yearly_sex_gender_plot + theme_bw()
```

> We can improve the plot by coloring by sex instead of species (species are
> already in separate plots, so we don't need to distinguish them further).

Note that even though the `color` and `group` aesthetic mappings are set, I can
*redefine them* via the `+` operator.

```{r facet-by-species-and-sex-colored}
yearly_sex_gender_plot + aes(color=sex, group=sex) + theme_bw()
```

> - plot average weight of each species over the course of the years

```{r average-weight-timeseries}
yearly_weight <- surveys_complete %>%
                 group_by(year, species_id, sex) %>%
                 summarise(avg_weight = mean(weight, na.rm = TRUE))
yearly_weight_plot <- ggplot(data = yearly_weight) + 
    aes(x=year, y=avg_weight, color = species_id, group = species_id) +
    geom_line() + theme_bw()
yearly_weight_plot
```

> - why do you think we see those steps in the plot?

The `facet_wrap` geometry extracts plots into an arbitrary number of dimensions
to allow them to cleanly fit on one page. On the other hand, the `facet_grid`
geometry allows you to explicitly specify how you want your plots to be
arranged via formula notation (`rows ~ columns`; a `.` can be used as
a placeholder that indicates only one row or column). 

> - make separate plots per sex since weight of males and females can differ a lot

```{r average-weight-time-facet_sex_rows}
# One column, facet by rows 
yearly_weight_plot + facet_grid(sex ~ .)
```

```{r average-weight-time-facet_sex_columns}
# One row, facet by column
yearly_weight_plot + facet_grid(. ~ sex)
```

So far our result looks quite good, but is far from being publishable.
What are other ways one can improve it? Take a look at the ggplot2 cheat sheet (https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf),
and write down at least three more ideas (you can leave them as comments
to the Etherpad).

Now, let's change names of axes to something more informative than 'year'
and 'n' and add a title to this figure:

```{r number_species_year_with_right_labels}
plt_with_labels <- ggplot(data = yearly_sex_counts) + 
    aes(x = year, y = n, color = sex, group = sex) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = 'Observed species in time',
        x = 'Year of observation',
        y = 'Number of species') + theme_bw()

plt_with_labels
```

Thanks to our efforts, the axes have much more informative names,
but are still quite small and hard to read.
Let's change their size (and font just for fun):

```{r number_species_year_with_right_labels_xfont_size}
plt_with_labels + theme(text=element_text(size=16, family="Arial"))

```

After our manipulations we notice that the values on the x-axis are still not
properly readable.  Let's change the orientation of the labels and adjust them
vertically and horizontally so they don’t overlap.  You can use a 90 degree
angle, or experiment to find the appropriate angle for diagonally oriented
labels.

```{r number_species_year_with_right_labels_xfont_orientation}
arial_grey_theme <- theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
                        axis.text.y = element_text(colour="grey20", size=12),
                        text=element_text(size=16, family="Arial"))
plt_with_labels + arial_grey_theme
```

Note that themes (and any other individual components of ggplot-based objects)
can be saved to variables for convenient access, as the previous code block
demonstrates. A useful application of this is that you can define a global
theme that contains text fonts and sizes, plot margins, etc. that you'd like to
apply to all of your plots and then selectively modify this theme for
individual plots without having to type out every setting each time. 

```{r, eval=FALSE}
# Setup plots
global_theme <- theme(...)
plot1_theme <- theme(...)
plot2_theme <- theme(...)
plot_template <- ggplot() + aes(...)
plot1_aes <- aes(...)   ## Additional custom aesthetic mappings for plot1
plot2_aes <- aes(...)   ## Additional custom aesthetic mappings for plot2
plot1 <- plot_template + plot1_aes + geom_point(...) + global_theme + plot1_theme
plot2 <- plot_template + plot2_aes + geom_line(...) + global_theme + plot2_theme

# Draw plots
plot1
plot2
```

Be aware that when overwriting settings via the `+` operator, order matters.
That is, `global_theme + plot1_theme` will overwrite any settings in
`global_theme` with those in `plot1_theme`, whereas `plot1_theme + global_theme` 
will overwrite any settings in `plot1_theme` with those in `global_theme`. 
This only applies to settings that you set, however -- if the `plot1` and
`global` themes don't share any common settings (i.e. are mutually exclusive),
neither will be affected by the other. 

With all of this information in hand, please take another five minutes to
either improve one of the plots generated in this exercise or create
a beautiful graph of your own. Use the ggplot2 cheat sheet, which we linked
earlier for inspiration.

Here are some ideas:

* See if you can change thickness of the lines.
* Can you find a way to change the name of the legend?
What about its labels?
* Use different color palette to improve the look
(http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)

When you created your perfect plot. You can save it to the file in your favourite format.
You can easily change the size by specifying the width and height of your plot.

```{r number_species_year_with_right_labels_xfont_save, eval=FALSE}
your_plot_here      # Draw the plot to the screen
ggsave("name_of_file.png", width=15, height=10)
```

Enjoy plotting with ggplot2!
