```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

# Data Carpentry R materials - Plotting

--------------------------------------------------

# Objectives

* generate plot from the data using the average weights of the animals as an
  example

--------------------------------------------------

```{r, echo=FALSE, purl=TRUE}
## Plotting
```

The mathematician Richard Hamming once said, "The purpose of computing is
insight, not numbers", and the best way to develop insight is often to visualize
data. Visualization deserves an entire lecture (or course) of its own, but we
can explore a few features of R's base plotting package.

Let's use the `surveys_summary` data that we generated and plot it.

R has built in plotting functions.

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt)
```

The axis labels are too big though, so you can't see them all. Let's change that.

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt, cex.names=0.4)
```

Alternatively, we may want to flip the axes to have more room for the species names:

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt, horiz=TRUE, las=1)
```

Let's also add some colors, and add a main title, label the axis:

```{r, purl=FALSE}
barplot(surveys_summary$mean_wgt, horiz=TRUE, las=1,
        col=c("lavender", "lightblue"), xlab="Weight (g)",
        main="Mean weight per species")
```

### Challenge

1. Create a new plot showing the standard deviation for each species. Choose
   one or more colors from
   [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf). (If you
   prefer, you can also specify colors using their hexadecimal values
   `#RRGGBB`.)

```{r, echo=FALSE, purl=TRUE}
## 1. Create a new plot showing the standard deviation for each species.
```

## More about plotting

There are lots of different ways to plot things. You can do `plot(object)` for
most classes included in R base. To explore some of the possibilities:

```{r, eval=FALSE, purl=FALSE}
?barplot
?boxplot
?plot.default
example(barplot)
```

There's also a plotting package called [`ggplot2`](http://docs.ggplot2.org/)
that adds a lot of functionality. The syntax takes some getting used to but it's
extremely powerful and flexible.

If you wanted to output this plot to a pdf file rather than to the screen, you
can specify where you want the plot to go with the `pdf()` function. If you
wanted it to be a JPG, you would use the function `jpeg()` (other formats
available: svg, png, ps).

Be sure to add `dev.off()` at the end to finalize the file. For `pdf()`, you can
create multiple pages to your file, by generating multiple plots before calling
`dev.off()`.


```{r, eval=FALSE, purl=TRUE}
pdf("mean_per_species.pdf")
barplot(surveys_summary$mean_wgt, horiz=TRUE, las=1,
        col=c("lavender", "lightblue"), xlab="Weight (g)",
        main="Mean weight per species")
dev.off()
```
