---
title: "SQL databases and R"
author: Data Carpentry contributors
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE, 
  echo = TRUE,
  purl = FALSE
)
```

```{r source, include=FALSE}
source("setup.R")
if (file.exists("portalR.sqlite")) file.remove("portalR.sqlite")
```

> ## Learning Objectives
>
>  By the end of this lesson the learner will know how to:
>
> * Access a database from R
> * Run SQL queries in R using RSQLite and dplyr.
> * Create an SQLite database from existing .csv files.

------------

# Introduction

So far, we have dealt with small datasets that easily fit into your computer's
memory. But what about datasets that are too large for your computer to
handle as a whole? In this case, storing the data outside of R and organizing it
in a database is helpful: Connecting to the database allows you to retrieve only
the chunks needed for the current analysis.

Even better, many large datasets are already available in public or private
databases. You can query them without having to download the data first.

R can connect to almost any existing database type. For example, the popular
[dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) package you
used in the previous lesson supports connecting to the popular open source
databases [sqlite](https://sqlite.org/), [mysql](https://www.mysql.com/) and
[postgresql](https://www.postgresql.org/), as well as google’s
[bigquery](https://cloud.google.com/bigquery/).

# The portal_mammals database

We will continue to explore the `survey` data you are
already familiar with from previous lessons.

The SQLite database is contained in a single file `portal_mammals.sqlite`
that you generated during the SQL lesson. If you don't have it, you can download
it from Figshare into the `data` subdirectory using:

```{r download, eval = FALSE}
dir.create("data", showWarnings = FALSE)
download.file(url = "https://ndownloader.figshare.com/files/2292171",
              destfile = "data/portal_mammals.sqlite")
```

# Connecting to databases with dplyr

We can point R to this database with `dplyr's` `src_sqlite()` command.

```{r connect}
library(dplyr)
mammals <- src_sqlite("data/portal_mammals.sqlite")
```

The `src_sqlite()` command does not load the data into the R session (as the 
`read.csv()` function did). Instead, it merely instructs R to connect to
the `SQLite` database contained in the `portal_mammals.sqlite` file.

(You can use the `src_mysql()`, `src_postgres()` and `src_bigquery()` to connect
to the other database types supported by `dplyr`.)

Let's take a closer look at the `mammals` database we just connected to

```{r tables, results='markup'}
mammals
```

Just like a spreadsheet with multiple worksheets, a SQLite database can contain
multiple tables. In this case three of them are listed in the `tbls` row in the output above:

* plots
* species,
* surveys

Let's explore these tables next!

## Querying the database

We specify the table of interest with `dplyr's` `tbl()` function. First, we focus 
on the `surveys` table.

```{r table_details, results='markup'}
surveys <- tbl(mammals, "surveys")
head(surveys, n = 10)
```

This output of the `head` command looks just like a regular `data.frame`:
The table has 9 columns and the `head()` command shows us the first 10 rows. 

Let's check how many rows there are in total:

```{r nrows, results='markup'}
nrow(tbl)
```

That's strange - R doesn't know how many rows the `survey` table contains - it
returns `NULL` instead. 

The reason for this behavior highlights a key difference between using `dplyr`
on datasets in memory (e.g. loaded into your R session via `read.csv()`) and
those provided by a database. To understand it, we take a closer look at how
`dplyr` communicates with our SQLite database.

## SQL translation

Relational databases typically use a special-purpose language,
[Structured Query Language (SQL)](https://en.wikipedia.org/wiki/SQL),
to manage and query data.

For example, the following SQL query returns the first 10 rows from the
`surveys` table:

```
SELECT *
FROM `surveys`
LIMIT 10
```

Behind the scenes, `dplyr`

1. translates your R code into SQL,
2. submits it to the database
3. translates the database's response into an R data.frame

To lift the curtain, we can use `dplyr's` `explain()` function to show which SQL
commands are actually sent to the database:

```{r explain, message=TRUE, results='hide'}
explain(head(surveys, n = 10))
```

The first part of the output shows the actual SQL query sent to the database;
it matches our manually constructed `SELECT` statement above. 

Instead of having to formulate the SQL query ourselves - and
having to mentally switch back and forth between R and SQL syntax - we can
delegate this translation to `dplyr`. (You don't even need to know SQL to interact
with a database via `dplyr`!)

`dplyr`, in turn, doesn't do the real work of subsetting the table, either.
Instead, it merely sends the query to the database, waits for its response and
returns it to us.

That way, R never get's to see the full `surveys` table - and that's why it could
tell us how many rows it contains. On the bright side, this allows us to work
with large datasets - even too large to fit into our computer's memory.

`dplyr` can translate many different query types into SQL allowing us e.g.
to `select()` specific columns, `filter()` rows or join tables. 

Too see this in action, let's compose a few queries with `dplyr`.

## Simple database queries

First, let's only request rows of the `surveys` table in which `weight` is less
than 5 and keep only the species_id, sex, and weight columns.

```{r pipe, results='markup'}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

Executing this command will return a table with 10 rows and the requested
`species_id`, `sex` and `weight` columns. Great!

... but wait, why are there only 10 rows?

The last line:

```
# ... with more rows
```

indicates that there are more results that fit our filtering criterion. Why was
R lazy and only retrieved 10 of them?

## Lazyness

Hadley Wickham, the author of `dplyr`
[explains](https://cran.r-project.org/web/packages/dplyr/vignettes/databases.html):

> When working with databases, dplyr tries to be as lazy as possible:
> 
> * It never pulls data into R unless you explicitly ask for it.
> * It delays doing any work until the last possible moment: it collects together
> everything you want to do and then sends it to the database in one step.

When you construct a `dplyr` query, you can connect multiple verbs into a single
pipeline. For example, we combined the `filter()` and `select()` verbs using the
`%>%` pipe.

If we wanted to, we could add on even more steps, e.g. remove the `sex` column
in an additional `select` call:

```{r pipe2, results='markup'}
data.subset <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

data.subset %>%
  select(-sex)
```

Just like the first `select(species_id, sex, weight)` call, the `select(-sex)`
command is not executed by R. It is sent to the database instead. Only the
_final_ result is retrieved and displayed to you.

Of course, we could always add on more steps, e.g. we could filter by
`species_id` or  minimum `weight`. That's why R doesn't retrieve the full set 
of results - instead it only retrieves the first 10 results from the database
by default. (After all, you might want to add an additional step and get the
database to do more work...)

To instruct R to stop being lazy, e.g. to retrieve all of the query results from
the database, we add the `collect()` command to our pipe. It indicates that our
database query is finished: time to get the _final_ results and load them into
the R session.

```{r collect, results='markup'}
data.subset <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight) %>%
  collect()
```

Now we have all 17 rows that match our query in a `data.frame` and can continue
to work with them exclusively in R, without communicating with the database.

## Complex database queries

`dplyr` enables database queries across one or multiple database tables, using
the same single- and multipe-table verbs you encountered previously. (This means
you can use the same commands regardless of whether you interact with a remote
database or local dataset!)

For example, the `plots` table in the database contains information about the
different plots surveyed by the reseachers. To access it, we point the `tbl()`
command to it:

```{r plots, results='markup'}
plots <- tbl(mammals, "plots")
plots
```

The `plot_id` column also features in the `surveys` table.

```{r surveys, results='markup'}
surveys
```

Because `plot_id` is listed in both tables, we can use it to look up matching
records. (Combining records from different tables is called a `join`.)

For example, let's extract all surveys for the first plot, which has `plot_id` 1:

```{r join, results='markup'}
plots %>%
  filter(plot_id == 1) %>%
  inner_join(surveys) %>% 
  collect()
```

**Note:** Without the `collect()` statement, only the first 10 matching rows
are returned. By adding `collect()`, the full set of 1,985 is retrieved.

> ## Challenge 1
>
>  Write a query that returns the number of rodents observed in each plot in 
>  each year.
>
>  Hint: Connect to the species table and write a query that joins the species
>  and survey tables together.
>  The query should return counts of rodents by year.

```{r left_join, results='markup'}
species <- tbl(mammals, "species")

left_join(surveys, species) %>%
  filter(taxa == "Rodent") %>%
  group_by(taxa, year) %>%
  tally %>%
  collect()
```

> ## Challenge 2
>
>  Write a query that returns counts of genus by `plot_id`
>
>  Hint: Write a query that joins the species, plot, and survey tables together.
>  The query should return counts of genus by plot type.

```{r genus_by_type, results='markup'}
genus.counts <- left_join(surveys, plots) %>%
  left_join(species) %>%
  group_by(plot_type, genus) %>%
  tally %>%
  collect()
```

## Creating a new SQLite database

So far, we have used a previously prepared SQLite database. But we can also
use R to create a new database, e.g. from existing `csv` files.  Let's recreate 
the mammals database that we've been working with, in R. First let's read in the
`csv` files.

```{r data_frames}
species <- read.csv("data/species.csv")
surveys <- read.csv("data/surveys.csv")
plots <- read.csv("data/plots.csv")
```

Creating a new SQLite database with `dplyr` is easy. You can re-use the same
command we used above to open an existing `.sqlite` file. The `create = TRUE`
argument instructs R to create a new, empty database instead.

**Caution:** When `create = TRUE` is added, any existing database at the same
location is overwritten _without warning_.

```{r create_database}
myDB <- "portalR.sqlite"
my_db <- src_sqlite(myDB, create = TRUE)
```

Currently, our new databse is empty, it doesn't contain any tables:

```{r empty, results='markup'}
my_db
```

To add tables, we copy the existing data.frames into the database one by one:

```{r copy, results='hide'}
copy_to(my_db, surveys)
copy_to(my_db, plots)
my_db
```

If you check the location of our database you'll see that data is automatically
being written to disk. R and `dplyr` not only provide easy ways to query
existing databases, they also allows you to easily create your own databases
from flat files!

> ## Challenge 3
>
> Add the remaining species table to the my_db database and run some of your
> queries from earlier in the lesson to verify that you have
> faithfully recreated the mammals database.

**Note:** In this example, we first loaded all of the data into the R session by
reading the three csv files. Because all the data has to flow through R, 
this is not suitable for very large datasets.

# Connecting with RSQlite

`dplyr` is a powerful tool focussed on retrieving and analyzing datasets by
generating `SELECT` SQL statements, but it doesn't modify the database itself.
For example, dplyr does not offer functions to `UPDATE` or `DELETE` entries.

Luckily, R can also communicate with your database directly. For example, the
`RSQLite` R package offers functions to send, retrieve and modify SQLite
databases.

Similarly, you should be able to connect to almost any database in R via
[JDBC](http://cran.r-project.org/web/packages/RJDBC/index.html)
or [ODBC](http://cran.r-project.org/web/packages/RODBC/index.html), or specific
database packages (such as we are doing, or
[MySQL](http://cran.r-project.org/web/packages/RMySQL/index.html) ).

Let's connect to our database using `RSQLite` and repeat our queries - this time
using standard SQL statements. (We will need the RSQLite package, so be sure to
install it first via `install.packages('RSQLite')`.)

First, we create a connection to the `portal_mammals.sqlite` database, just like
we did with `dplyr's` `src_sqlite` function above.

```{r}
library(RSQLite)
conn <- dbConnect(drv = SQLite(), dbname= "data/portal_mammals.sqlite")
```

## db helper functions

Now we can use helper functions from `RSQLite`, e.g. to list both tables and 
fields within a table.

```{r table details, purl=FALSE}
dbListTables(conn)
dbListFields(conn, "surveys")
```

The `RSQLite` R package comes with additional helper functions e.g. to create
new databases, write tables, retrieve results in small chunks or load entire
tables into your R session. Check its
[vignette](https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html)
for more examples!

## Executing arbitrary SQL queries

In addition to these helper functions, we can submit _any_ valid SQL
query with the `dbGetQuery` function. For example, let's count the number of
rows in the `surveys` table:

```{r}
dbGetQuery(conn,"SELECT count(*) FROM surveys")
```

## Complex SQL queries

Of course, we can also issue more complex queries. Let's revisit 
__Challenge 1__ from above, but this time we will compose the SQL statement
ourselves, without `dplyr's` help.

> ## Challenge 1, redux
>
>  Write a query that returns the number of rodents observed in each plot in 
>  each year.
>  Hint: You can join multiple tables together in SQL using the following syntax
>  where foreign key refers to your unique id (e.g., `species_id`):
>
>     SELECT table.col, table.col
>     FROM table1 JOIN table2  
>     ON table1.key = table2.key  
>     JOIN table3 ON table2.key = table3.key

First, we write our SQL query and store in a variable, e.g. `query`:

```{r genus by type, purl=FALSE}
query <- paste("
SELECT a.year, b.taxa,count(*) as count
FROM surveys a
JOIN species b
ON a.species_id = b.species_id
AND b.taxa = 'Rodent'
GROUP BY a.year, b.taxa",
sep = "" )
```

Then, we submit this query to the database and collect its answer in the
`result` object:

```{r}
result <- dbGetQuery(conn, query)
head(result)
```

* Does the result match the output from our `dplyr` solution above?
* Which solution do you find easier to read and understand?

The `dbGetQuery` function accepts arbitrary SQL statements, opening up the full
functionality of the SQLite database to us!

## Cleaning up

It is good practice to always close a connection that you open in R. Let's do
that next. Note that once you've closed a connection, you will have to open
a new connection to query and import the data again.

```{r close Connection, purl=FALSE}
dbDisconnect(conn)
print(conn)
```

