```{r, echo=FALSE}
opts_chunk$set(results='hide')
```
# Taking notes and creating reports:

## Introducing Knitr

`Knitr` is an R package that makes your code neat, pretty, and shows your notes, code, and output simultaneously in an `html` document. You create these documents in `.Rmd` files. You can write in `LateX` or `md`. 

knitr extracts R code in the input document (.Rmd), evaluates it and writes the results to the output document (html). There are two types of R code: chunks (code as separate paragraphs) and inline R code.

```{r, eval=FALSE}
install.packages("knitr")
```	
```{r}
library(knitr)
```    

Restart RStudio. 
Open a new `Rmd` file. 

In Rmd, anything between lines that start and end with triple quotes ``` will be run as R code.

```{r}
summary(cars)
```

### Challenge

1. Open an new .Rmd script and save it as inflammation_report.Rmd
2. Copy and paste the code as embedded R chunks to read in the data and plot average inflammation, or the heat map that we created.
3. Add a few notes describing what the code does and what the main findings are.
4. `KNIT` and view the html
