---
layout: topic
title: tidyr
author: Data Carpentry contributors
---


```{r, echo=FALSE, message = FALSE}
setwd("/Users/kivan/repos/R-ecology-lesson")
```

```{r, echo=FALSE, purl=FALSE, message = FALSE, warning=FALSE}
source("setup.R")
surveys <- read.csv("data/portal_data_joined.csv")
```

------------

# Defining data structure with tidyr

Tidying messy datasets

Real datasets can, and often do, violate the three precepts of tidy data in almost every way imaginable. While occasionally you do get a dataset that you can start analysing immediately, this is the exception, not the rule. This section describes the five most common problems with messy datasets, along with their remedies:

* Column headers are values, not variable names.
* Multiple variables are stored in one column.
* Variables are stored in both rows and columns.
* Multiple types of observational units are stored in the same table.
* A single observational unit is stored in multiple tables.

```{r, eval = FALSE, purl = FALSE}
install.packages("tidyr")
```


```{r, message = FALSE, purl = FALSE}
library("tidyr")    ## load the package
```

## What is `tidyr`?

The package `tidyr` 

This addresses a common problem with R in that all operations are conducted in
memory and thus the amount of data you can work with is limited by available
memory. The database connections essentially remove that limitation in that you
can have a database of many 100s GB, conduct queries on it directly, and pull
back just what you need for analysis in R.

To learn more about `dplyr` after the workshop, you may want to check out this
[handy tidyr cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).


## Column headers are values, not variable names
