---
layout: topic
title: Tidy Data - A foundation for wrangling in R
author: Data Carpentry contributors
---


```{r, echo=FALSE, message = FALSE}
setwd("/Users/kivan/repos/R-ecology-lesson")
```

```{r, echo=FALSE, purl=FALSE, message = FALSE, warning=FALSE}
source("setup.R")
surveys <- read.csv("data/portal_data_joined.csv")
```

------------

# Defining data structure with tidyr

Tidying messy datasets

Real datasets can, and often do, violate the three precepts of tidy data in almost every way imaginable. While occasionally you do get a dataset that you can start analysing immediately, this is the exception, not the rule. This section describes the five most common problems with messy datasets, along with their remedies:

* Column headers are values, not variable names.
* Multiple variables are stored in one column.
* Variables are stored in both rows and columns.
* Multiple types of observational units are stored in the same table.
* A single observational unit is stored in multiple tables.

```{r, eval = FALSE, purl = FALSE}
install.packages("tidyr")
```


```{r, message = FALSE, purl = FALSE}
library("tidyr")    ## load the package
```

## What is `tidyr`? ##

The package `tidyr` 



To learn more about `tidyr` after the workshop, you may want to check out this
[handy tidyr cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).

## Data semantics

A dataset is a collection of values, usually either numbers (if quantitative) or strings (if qualitative). Values are organised in two ways. Every value belongs to a variable and an observation. A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units. An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes.

A tidy version of the pregnancy data looks like this: (you’ll learn how the functions work a little later)

## Column headers are values, not variable names
