# R Objects / Data Structure

R has many different structures for storing the data you want to
analyze. The most commonly used are

- vectors,
- lists,
- matrices,
- arrays, and 
- data frames.

It can be determined with the function `class()`.

The *type* or *mode* of an object defines how it is stored. It could
be

- a character,
- a numeric value,
- an integer,
- a complex number, or
- a logical value (Boolean value: TRUE/FALSE).

It can be determined with the function `mode()`.

Below is a short definition  and quick example of each of these data
structures which you can use to decide which is the best for
representing your data  in R.

## Vectors
It is the most basic data type. Vectors only have one dimension and
all their elements must be the same mode. There are various ways to
create vectors. The simplest one is with the `c` function.

```{r}
(v1 <- c(1, 2, 5))
```

The `c` function allows to combine its arguments. If the arguments
are of various modes, they will be reduced to their lowest common
type:

```{r}
(v2 <- c(1, 3, "a"))
mode(v2)
```

Objects can be explicitly coerced with the `as.` function.

```{r}
as.character(v1)
```

You can also use the `:` operator or the `seq` function:

```{r}
1:10
seq(from = 5, to = 25, by = 5)
```

### Factors
Factor is a special type of vector to store categorical values.

```{r}
(breed <- factor(c("Holstein", "Holstein", "Brown Swiss", "Holstein",
                  "Ayrshire", "Canadian", "Canadian", "Brown Swiss",
                  "Holstein", "Brown Swiss", "Holstein")))
```
It stores values as a set of labeled integers. Some functions treat
factors differently from numeric vectors.

```{r}
table(breed)
```

## Lists
A list is an ordered collection of objects where the objects can be of
different modes.

```{r}
(l <- list("a", "b", "c"))
```

Each element of a list can be given a name and referred to by that
name. Elements of a list can be accessed by their number or their name.

```{r}
(cow <- list(breed = "Holstein", age = 3, last.prod = c(25, 35, 32)))
cow$breed
cow[[1]]
```

Lists can be used to hold together multiple values returned from a
function. For example the elements used to create an histogram can be
saved and returned:

```{r}
h <- hist(islands)
str(h)
```

The function `str` is used here. It stands for *structure* and shows
the internal structure of an R object.

## Matrices
A matrix is a rectangular array of numbers. Technically, it is a
vector with two additional attributes: number of rows and number of
columns. This is what you would use to analyze a spreadsheet full of
only numbers, or only words. You can create a matrix with the `matrix`
function:

```{r}
(m <- rbind(c(1, 4), c(2, 2)))
(m <- matrix(data = 1:12, nrow = 4, ncol = 3,
            dimnames = list(c("cow1", "cow2", "cow3", "cow4"),
                c("milk", "fat", "prot"))))
```

## Arrays
If a matrix is a two-dimensional data structure, we can add *layers*
to the data and have further dimensions in addition to rows and
columns. These datasets would be arrays. It can be
created with the `array` function:

```{r}
(a <- array(data = 1:24, dim = c(3, 4, 2)))
```

## Data frames
Data frames are used to store tabular data: multiple rows, columns and
format.

```{r}
(df <- data.frame(cow = c("Moo-Moo", "Daisy", "Elsie"),
                 prod = c(35, 40, 28),
                 pregnant = c(TRUE, FALSE, TRUE)))
```
