---
title: ggplot2 essentials (20 min)
teaching: 20
exercises: 10
source: Rmd
---

::::::::::::::::::::::::::::::::::::::: objectives

- Use ggplot2 to generate plots.
- Know the three components to a basic ggplot2: data, aesthetic mappings, geometric objects.
- Manipulate the aesthetic mappings of a plot.
- Save a plot created with ggplot() to disk.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How do I create plots with `ggplot2`?
- How do I save the plots I created?

::::::::::::::::::::::::::::::::::::::::::::::::::

Plotting is a key component of exploratory data analysis and a powerful way to identify patterns, trends, and relationships between variables in your dataset.

In this lesson, we will use the `ggplot2` package, a widely used system for creating clear and flexible data visualisations in R.

`ggplot2` is built on the grammar of graphics, which is the idea that any plot can be built from the same set of components: a **data set**, **mapping aesthetics**, and graphical **layers**:

- **Data** are the data that you, the user, provide.

- **Aesthetic Mappings ** are what connect the data to the graphics. They tell `ggplot()` how to use your data to affect how the graph looks, such as changing what is plotted on the X or Y axis, or the size or color of different data points.

- **Geometric objects (geoms)** are the parts of the plot that we can see, such as points, lines, or bars. Each **geom** creates a different type of plot (e.g. scatter plots, histograms, bar charts). In `ggplot2`, plots are built up in layers, with each geom added as new layer.

## Data

In this lesson we will use the `mpg` dataset, which is included with the `ggplot2` package. This dataset contains fuel economy data for a range of cars. Fuel efficiency is recorded in miles per gallon (mpg), which is commonly used in the United States (in Australia, we more often use litres per 100 km). Higher values indicate better fuel efficiency.

Because `mpg` is built into `ggplot2`, we can use it directly without needing to load an external file.

```r
library(ggplot2)
str(mpg)
```
```output
tibble [234 × 11] (S3: tbl_df/tbl/data.frame)
 $ manufacturer: chr [1:234] "audi" "audi" "audi" "audi" ...
 $ model       : chr [1:234] "a4" "a4" "a4" "a4" ...
 $ displ       : num [1:234] 1.8 1.8 2 2 2.8 2.8 3.1 1.8 1.8 2 ...
 $ year        : int [1:234] 1999 1999 2008 2008 1999 1999 2008 1999 1999 2008 ...
 $ cyl         : int [1:234] 4 4 4 4 6 6 6 4 4 4 ...
 $ trans       : chr [1:234] "auto(l5)" "manual(m5)" "manual(m6)" "auto(av)" ...
 $ drv         : chr [1:234] "f" "f" "f" "f" ...
 $ cty         : int [1:234] 18 21 20 21 16 18 18 18 16 20 ...
 $ hwy         : int [1:234] 29 29 31 30 26 26 27 26 25 28 ...
 $ fl          : chr [1:234] "p" "p" "p" "p" ...
 $ class       : chr [1:234] "compact" "compact" "compact" "compact" ...
```

### Data Dictionary

The main variables we will use in this lesson are:

- displ: engine size (litres)
- hwy: highway fuel efficiency (miles per gallon)
- cty: city fuel efficiency (miles per gallon)
- class: type of car (e.g. SUV, compact)
- drv: drive type (front-wheel, rear-wheel, 4-wheel)

Using this information, which variables might you choose to explore a relationship between engine size and fuel efficiency?

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 1: Know your data

Before creating any plots, it’s useful to briefly explore the dataset to understand what variables are available and what they represent.

Take a minute to explore the dataset:

- What types of variables are included (numeric, categorical)?
- Which variable might you use for the **x-axis** if you wanted to explore engine size?
- Which variable could you use for the **y-axis** to represent fuel efficiency?
- Is there a categorical variable you could use to group points?

:::::::::::::::  solution

- numeric: displ, year, cyl, cty, hwy
- categorical: manufacturer, model, trans, drv, fl, class
- engine size: displacement
- fuel efficiency: mpg
- any of the categorical variables can be used to group points

:::::::::::::::::::::::::
:::::::::::::::::::::::::::::::::::::::

Let’s start off building an example using the `mgp` data. The most basic function is `ggplot()`, which lets R know that we’re creating a new plot. Any of the arguments we give the `ggplot()` function are the _global_ options for the plot - they apply to all layers on the plot.

```{r blank-ggplot, message=FALSE, fig.alt="Blank plot, before adding any mapping aesthetics to ggplot()."}
ggplot(data = mpg)
```

Here we called `ggplot()` and told it what data we want to show on our plot. This is not enough information to actually draw anything. However, it does create a blank plot that helps demonstrate how the components are put together. We're essentially providing the base layer for other elements to be added on to.

Next we're going to add in the **mapping aesthetics** using the `aes()` function. `aes()` tells `ggplot()` how variables in the **data** map to *aesthetic* properties of the plot, such as which columns of the data should be used for the **x** and **y** locations.

```{r ggplot-with-aes, message=FALSE, fig.alt="Plotting area with axes for a scatter plot of displacement and hwy fuel efficiency but no data points visible."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy))
```

Here we told `ggplot()` we want to plot the "displ" column of the data frame on the x-axis, and the "hwy" column on the y-axis. Notice that we didn't need to explicitly pass `aes` these columns (e.g. `x = mpg[, "disp"]`). This is because
`ggplot()` is designed to look in the **data** for that column!

Notice again that we still don't have a plot. The third and final component needed to make a plot is a **geom** function to tell `ggplot()` how to visually represent the data. For example, if we want to use **points** to represent the data in our plot, we use the `geom_point()` function.

The `+` is used to add layers to a plot.

```{r ggplot-with-aes-geom, message=FALSE, fig.alt="Scatter plot of displacement vs hwy fuel efficiency with data points."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

There we have it! A scatter plot of points that represents the relationship between **x** ('displacement') and **y** ('highway fuel efficiency') in our dataset. 

These three components — **data**, **aesthetic mappings**, and **geoms** — are all you need to create a basic plot in `ggplot2`.

In the rest of this lesson, we will build on this foundation to create different types of visualisations and explore how to adapt them for different questions.

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 2 Aesthetic Mappings

Modify the example so that the plot shows how city fuel efficiency relates to engine displacement.

What can we say about the relationship between city fuel efficiency and displacement?

:::::::::::::::  solution

```{r ggplot-with-aes-geom-cty, eval=FALSE, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points."}}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty )) + 
  geom_point()
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 3 Aesthetic Mappings - add colour

In the previous examples we've used the `aes` function to tell the scatter plot **geom** about the **x** and **y** locations of each point.

Another *aesthetic* property we can modify is the *colour* of the points.

Modify the code from the previous challenge to **colour** the points by the "year"
column. 

- What trends do you see in the data? 
- Are they what you expected?

:::::::::::::::  solution

The solution presented below adds the argument `colour = year` to the call of the `aes()` function. 

The general trend seems to indicate a decrease in fuel efficiency as engine displacement increases. Cars built in year 2008 appear to have higher fuel efficiency overall than cars.

:::::::::::::::::::::::::
  
::::::::::::::::::::::::::::::::::::::::::::::::::

## Saving the plot

The `ggsave()` function allows you to save a plot created with `ggplot2` quickly and easily with just a filename:

```{r save}
ggsave(filename = "my_ggplot.png")
```

Notice that we did not provide any additional arguments to `ggsave()`.

- If we omit the `plot` argument, `ggsave()` will automatically save the last plot you created with `ggplot`.
- If we omit the `device` argument, ggsave() will use the file extension to determine the device.

In this episode, we introduced the grammar of graphics and the core components of ggplot2: data, aesthetic mappings, and geometric layers. Using these components, we created our first plot with `geom_point()` and explored how variables can be mapped to visual properties such as colour. 

In the next episodes, we will build on these basics to explore our dataset in more detail. We will use a selection commonly used **geoms** to examine relationships between variables, the distribution of individual variables, differences between groups, and the composition of our data.


:::::::::::::::::::::::::::::::::::::::: keypoints

- `ggplot2` can quickly create simple plots for exploratory data analysis.
- `ggplot2` plots are built around the grammer of graphics where the key components of a plot can be extended with the additional **geom** layers.

::::::::::::::::::::::::::::::::::::::::::::::::::


