---
title: Working with data in the tidyverse
author: Data Carpentry contributors.& Nathaniel Porter
teaching: 20
exercises: 10
---

```{r, echo=FALSE, purl=FALSE, message=FALSE}
source("setup.R")
```

------------------------------------------------------------------------

::::::::::::::::::::::::::::::::::::: objectives

-   Load external data from a .csv file into a data frame.
-   Install and load packages.
-   Describe what a data frame is.
-   Summarize the contents of a data frame.
-   Wrangle data with **dplyr**.
-   Export a data frame to a .csv file.
::::::::::::::::::::::::::::::::::::::::::::::::

------------------------------------------------------------------------

## Loading the survey data

```{r, echo=FALSE, purl=TRUE}
### Loading the survey data
```

We are investigating the animal species diversity and weights found
within plots at our study site. The dataset is stored as a comma
separated value (CSV) file. Each row holds information for a single
animal, and the columns represent:

| Column          | Description                                  |
|-----------------|----------------------------------------------|
| record_id       | Unique id for the observation                |
| month           | month of observation                         |
| day             | day of observation                           |
| year            | year of observation                          |
| plot_id         | ID of a particular experimental plot of land |
| species_id      | 2-letter code                                |
| sex             | sex of animal ("M", "F")                     |
| hindfoot_length | length of the hindfoot in mm                 |
| weight          | weight of the animal in grams                |
| genus           | genus of animal                              |
| species         | species of animal                            |
| taxon           | e.g. Rodent, Reptile, Bird, Rabbit           |
| plot_type       | type of plot                                 |

### Downloading the data

We created the folder that will store the downloaded data (`data_raw`)
in the chapter ["Before we
start"](https://datacarpentry.org/R-ecology-lesson/00-before-we-start.html#Organizing_your_working_directory).
If you skipped that part, it may be a good idea to have a look now, to
make sure your working directory is set up properly.

We are going to use the R function `download.file()` to download the CSV
file that contains the survey data from Figshare, and we will use
`read_csv()` to load the content of the CSV file into R.

Inside the `download.file` command, the first entry is a character
string with the source URL
("<https://ndownloader.figshare.com/files/2292169>"). This source URL
downloads a CSV file from figshare. The text after the comma
("data_raw/portal_data_joined.csv") is the destination of the file on
your local machine. You'll need to have a folder on your machine called
"data_raw" where you'll download the file. So this command downloads a
file from Figshare, names it "portal_data_joined.csv" and adds it to a
preexisting folder named "data_raw".

```{r, eval=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "data_raw/portal_data_joined.csv")
```

### Reading the data into R

The file has now been downloaded to the destination you specified, but R
has not yet loaded the data from the file into memory. To do this, we
can use the `read_csv()` function from the **`readr``** package.

Packages in R are basically sets of additional functions that let you do
more stuff. The functions we've been using so far, like `round()`,
`sqrt()`, or `c()` come built into R. Packages give you access to
additional functions beyond base R. A similar function to `read_csv()`
from the tidyverse package is `read.csv()` from base R. We don't have
time to cover their differences but notice that the exact spelling
determines which function is used. Before you use a package for the
first time you need to install it on your machine, and then you should
import it in every subsequent R session when you need it.

The **`tidyverse`** package contains **`readr`** and other packages we'll use. To install it, we can type
`install.packages("tidyverse")` straight into the console. In fact, it's
better to write this in the console than in our script for any package,
as there's no need to re-install packages every time we run the script.
Then, to load the package type:

```{r, message=FALSE, purl=FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```

Now we can use the functions from any **`tidyverse`** package. Let's use
`read_csv()` to read the data into a data frame (we will learn more
about data frames later):

```{r, eval=TRUE, purl=FALSE}
surveys <- read_csv("data_raw/portal_data_joined.csv")
```

When you execute `read_csv` on a data file, it looks through the first
1000 rows of each column and guesses its data type. For example, in this
dataset, `read_csv()` reads `weight` as `col_double` (a numeric data
type), and `species` as `col_character`. You have the option to specify
the data type for a column manually by using the `col_types` argument in
`read_csv`.

::: callout
### Note

`read_csv()` is actually a special case of `read_delim()` that assumes  fields are delineated by commas. Check out the help by typing `?read_csv` to learn more.

You may sometimes see a similar function called `read.csv()` with a dot instead of an underscore. This is an older function that doesn't require **`tidyverse`** with different arguments and slightly different output.

:::

We can see the contents of the first few lines of the data by typing its
name: `surveys`. By default, this will show you as many rows and columns
of the data as fit on your screen. If you wanted the first 50 rows, you
could type `print(surveys, n = 50)`

We can also extract the first few lines of this data using the function
`head()`:

```{r, results="show", purl=FALSE}
head(surveys)
```

Unlike the `print()` function, `head()` **returns** the extracted data. You
could use it to assign the first 100 rows of `surveys` to an object
using `surveys_sample <- head(surveys, 100)`. This can be useful if you
want to try out complex computations on a subset of your data before you
apply them to the whole data set. There is a similar function that lets
you extract the last few lines of the data set. It is called (you might
have guessed it) `tail()`.

To open the dataset in RStudio's Data Viewer, use the `view()` function:

```{r, eval=FALSE, purl=FALSE}
view(surveys)
```

::: callout
### Note

There are two functions for viewing which are case-sensitive. Using
`view()` with a lowercase 'v' is part of tidyverse, whereas using
`View()` with an uppercase 'V' is loaded through base R in the `utils`
package.
:::

## What are data frames?

When we loaded the data into R, it got stored as an object of class
`tibble`, which is a special kind of data frame (the difference is not
important for our purposes, but you can learn more about tibbles
[here](https://tibble.tidyverse.org/)). Data frames are the *de facto*
data structure for most tabular data, and what we use for statistics and
plotting. Data frames can be created by hand, but most commonly they are
generated by functions like `read_csv()`; in other words, when importing
spreadsheets from your hard drive or the web.

A data frame is the representation of data in the format of a table
where the columns are vectors that all have the same length. Because
columns are vectors, each column must contain a single type of data
(e.g., characters, integers, factors). For example, here is a figure
depicting a data frame comprising a numeric, a character, and a logical
vector.

![](./fig/data-frame.svg)

We can see this also when inspecting the <b>str</b>ucture of a data
frame with the function `str()`:

```{r, purl=FALSE}
str(surveys)
```

## Inspecting data frames

We already saw how the functions `head()` and `str()` can be useful to
check the content and the structure of a data frame. Here is a
non-exhaustive list of functions to get a sense of the content/structure
of the data. Let's try them out!

-   Size:

    -   `dim(surveys)` - returns a vector with the number of rows in the
        first element, and the number of columns as the second element
        (the **dim**ensions of the object)
    -   `nrow(surveys)` - returns the number of rows
    -   `ncol(surveys)` - returns the number of columns

-   Content:

    -   `head(surveys)` - shows the first 6 rows
    -   `tail(surveys)` - shows the last 6 rows
    -   `glimpse(surveys)` - turns a data frame on its side to focus on the columns. Similar to `str()` but with nicer printing.

-   Names:

    -   `names(surveys)` - returns the column names (synonym of
        `colnames()` for `data.frame` objects)
    -   `rownames(surveys)` - returns the row names

-   Summary:

    -   `str(surveys)` - structure of the object and information about
        the class, length and content of each column
    -   `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic"; they can be used on other
types of objects besides `data.frame`.

:::::::::::::::::::::::::::::::::::::::::: callout

**`tidyverse` vs. base R**
The **`tidyverse`** package is an "umbrella-package" that installs **`dplyr`** and several other useful packages for data analysis, such as  **`ggplot2`**, **`tibble`**, etc.

As we begin to delve more deeply into the `tidyverse`, we should briefly pause to mention some of the reasons for focusing on the `tidyverse` set of tools. In R, there are often many ways to get a job done, and there are other approaches that can accomplish tasks similar to the `tidyverse`.

The phrase **base R** is used to refer to approaches that utilize functions contained in R's default packages. We have already used some base R functions, such as `str()`, `head()`, and `mean()`, and we will be using more scattered throughout this lesson. However, there are some key base R approaches we will not be teaching. These include square bracket subsetting and base plotting. You may come across code written by other people that looks like `surveys[1:10, 2]` or `plot(surveys$weight, surveys$hindfoot_length)`, which are base R commands. If you're interested in learning more about these approaches, you can check out other Carpentries lessons like the [Software Carpentry Programming with R](https://swcarpentry.github.io/r-novice-inflammation/) lesson.

We choose to teach the `tidyverse` set of packages because they share a similar syntax and philosophy, making them consistent and producing highly readable code. They are also very flexible and powerful, with a growing number of packages designed according to similar principles and to work well with the rest of the packages. The `tidyverse` packages tend to have very clear documentation and wide array of learning materials that tend to be written with novice users in mind. Finally, the `tidyverse` has only continued to grow, and has strong support from RStudio, which implies that these approaches will be relevant into the future.

::::::::::::::::::::::::::::::::::::::::::

## Data wrangling with dplyr

Next, we're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

## Selecting columns and filtering rows

To select columns of a data frame, use `select()`. The first argument
to this function is the data frame (`surveys`), and the subsequent
arguments are the columns to keep.

```{r, results="hide", purl=FALSE}
select(surveys, plot_id, species_id, weight)
```

To select all columns *except* certain ones, put a "-" in front of
the variable to exclude it.

```{r, results="hide", purl=FALSE}
select(surveys, -record_id, -species_id)
```

This will select all the variables in `surveys` except `record_id`
and `species_id`.

To choose rows based on a specific criterion, use `filter()`:

```{r, purl=FALSE}
filter(surveys, year == 1995)
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary data frame and use
that as input to the next function, like this:

```{r, purl=FALSE}
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)
```

This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl=FALSE}
surveys_sml <- select(filter(surveys, weight < 5), species_id, sex, weight)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then selecting).

The last option, pipes, are a recent addition to R. Pipes let you take the output of one function and send it directly to the next, which is useful when you need to do many things to the same dataset. There are two Pipes in R: 1) %>% (called magrittr pipe; made available via the magrittr package, installed automatically with dplyr) or 2) |> (called native R pipe and it comes preinstalled with R v4.1.0 onwards). Both the pipes are, by and large, function similarly with a few differences (For more information, check: https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/). The choice of which pipe to be used can be changed in the Global settings in R studio and once that is done, you can type the pipe with:

- <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> +
  <kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, purl=FALSE}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

In the above code, we use the pipe to send the `surveys` dataset first through
`filter()` to keep rows where `weight` is less than 5, then through `select()`
to keep only the `species_id`, `sex`, and `weight` columns. Since `%>%` takes
the object on its left and passes it as the first argument to the function on
its right, we don't need to explicitly include the data frame as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then." For instance,
in the example above, we took the data frame `surveys`, *then* we `filter`ed
for rows with `weight < 5`, *then* we `select`ed columns `species_id`, `sex`,
and `weight`. The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe we can accomplish
more complex manipulations of data frames.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl=FALSE}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```

Note that the final data frame is the leftmost part of this expression.

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge {.challenge}

Using pipes, subset the `surveys` data to include animals collected before
1995 and retain only the columns `year`, `sex`, and `weight`.

:::::::: solution

```{r, answer=TRUE, eval=FALSE, purl=FALSE}
surveys %>%
    filter(year < 1995) %>%
    select(year, sex, weight)
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Pipes Challenge:
##  Using pipes, subset the data to include animals collected
##  before 1995, and retain the columns `year`, `sex`, and `weight.`
```

## Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

To create a new column of weight in kg:

```{r, purl=FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000)
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, purl=FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)
```

If this runs off your screen and you just want to see the first few rows, you
can use a pipe to view the `head()` of the data. (Pipes work with non-**`dplyr`**
functions, too, as long as the **`dplyr`** or `magrittr` package is loaded).

```{r, purl=FALSE}
surveys %>%
  mutate(weight_kg = weight / 1000)
```

The first few rows of the output are full of `NA`s, so if we wanted to remove
those we could insert a `filter()` in the chain:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000)
```

`is.na()` is a function that determines whether something is an `NA`. The `!`
symbol negates the result, so we're asking for every row where weight *is not* an `NA`.

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge {.challenge}

Create a new data frame from the `surveys` data that meets the following
criteria: contains only the `species_id` column and a new column called
`hindfoot_cm` containing the `hindfoot_length` values (currently in mm)
converted to centimeters.
In this `hindfoot_cm` column, there are no `NA`s and all values are less
than 3.

**Hint**: think about how the commands should be ordered to produce this data frame!

:::::::: solution

```{r, answer=TRUE, eval=FALSE, purl=FALSE}
surveys_hindfoot_cm <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_cm = hindfoot_length / 10) %>%
    filter(hindfoot_cm < 3) %>%
    select(species_id, hindfoot_cm)
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Mutate Challenge:
##  Create a new data frame from the `surveys` data that meets the following
##  criteria: contains only the `species_id` column and a new column called
##  `hindfoot_cm` containing the `hindfoot_length` values converted to centimeters.
##  In this `hindfoot_cm` column, there are no `NA`s and all values are less
##  than 3.

##  Hint: think about how the commands should be ordered to produce this data frame!
```

## Split-apply-combine data analysis and the `summarize()` function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. Key functions of **`dplyr`** for this workflow are
`group_by()` and `summarize()`.

### The `group_by()` and `summarize()` functions

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the mean `weight` by sex:

```{r, purl=FALSE}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

You may also have noticed that the output from these calls doesn't run off the
screen anymore. It's one of the advantages of `tbl_df` over data frame.

You can also group by multiple columns:

```{r, purl=FALSE}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

Here, we used `tail()` to look at the last six rows of our summary. Before, we had
used `head()` to look at the first six rows. We can see that the `sex` column contains
`NA` values because some animals had escaped before their sex and body weights
could be determined. The resulting `mean_weight` column does not contain `NA` but
`NaN` (which refers to "Not a Number") because `mean()` was called on a vector of
`NA` values while at the same time setting `na.rm = TRUE`. To avoid this, we can
remove the missing values for weight before we attempt to calculate the summary
statistics on weight. Because the missing values are removed first, we can omit
`na.rm = TRUE` when computing the mean:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```

Here, again, the output from these calls doesn't run off the screen
anymore. If you want to display more data, you can use the `print()` function
at the end of your chain with the argument `n` specifying the number of rows to
display:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```

Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the minimum weight for each species for each sex:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
```

It is sometimes useful to rearrange the result of a query to inspect the values. For instance, we can sort on `min_weight` to put the lighter species first:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
```

To sort in descending order, we need to add the `desc()` function. If we want to sort the results by decreasing order of mean weight:

```{r, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
```

### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each sex, we would do:

```{r, purl=FALSE}
surveys %>%
  count(sex)
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarizing it by counting the number of observations in that group. In other words, `surveys %>% count()` is equivalent to:

```{r, purl=FALSE}
surveys %>%
  group_by(sex) %>%
  summarize(count = n())
```

For convenience, `count()` provides the `sort` argument:

```{r, purl=FALSE}
surveys %>%
  count(sex, sort = TRUE)
```

Previous example shows the use of `count()` to count the number of rows/observations
for *one* factor (i.e., `sex`).
If we wanted to count *combination of factors*, such as `sex` and `species`,
we would specify the first and the second factor as the arguments of `count()`:

```{r, purl=FALSE}
surveys %>%
  count(sex, species)
```

With the above code, we can proceed with `arrange()` to sort the table
according to a number of criteria so that we have a better comparison.
For instance, we might want to arrange the table above in (i) an alphabetical order of
the levels of the species and (ii) in descending order of the count:

```{r, purl=FALSE}
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))
```

From the table above, we may learn that, for instance, there are 75 observations of
the *albigula* species that are not specified for its sex (i.e. `NA`).

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge {.challenge}

1. How many animals were caught in each `plot_type` surveyed?

:::::::: solution

```{r, answer=TRUE, purl=FALSE}
surveys %>%
  count(plot_type)
```

:::::::::::::::::

2. Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
  length for each species (using `species_id`). Also add the number of
  observations (hint: see `?n`).

:::::::: solution

```{r, answer=TRUE, purl=FALSE}
surveys %>%
  filter(!is.na(hindfoot_length)) %>%
  group_by(species_id) %>%
  summarize(
      mean_hindfoot_length = mean(hindfoot_length),
      min_hindfoot_length = min(hindfoot_length),
      max_hindfoot_length = max(hindfoot_length),
      n = n()
    )
```

:::::::::::::::::

3. What was the heaviest animal measured in each year? Return the columns `year`,
  `genus`, `species_id`, and `weight`.

:::::::: solution

```{r, answer=TRUE, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(year) %>%
  filter(weight == max(weight)) %>%
  select(year, genus, species, weight) %>%
  arrange(year)
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Count Challenges:
##  1. How many animals were caught in each `plot_type` surveyed?

##  2. Use `group_by()` and `summarize()` to find the mean, min, and max
## hindfoot length for each species (using `species_id`). Also add the number of
## observations (hint: see `?n`).

##  3. What was the heaviest animal measured in each year? Return the
##  columns `year`, `genus`, `species_id`, and `weight`.
```

## Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new data sets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from data frames.

Before using `write_csv()`, we are going to create a new folder, `data`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data_raw` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data`
directory, so even if the files it contains are deleted, we can always
re-generate them.

In preparation for our next lesson on plotting, we are going to prepare a
cleaned up version of the data set that doesn't include any missing data.

Let's start by removing observations of animals for which `weight` and `hindfoot_length` are missing, or the `sex` has not been determined:

```{r, purl=FALSE}
surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex
```

Because we are interested in plotting how species abundances have changed
through time, we are also going to remove observations for rare species (i.e.,
that have been observed less than 50 times). We will do this in two steps: first
we are going to create a data set that counts how often each species has been
observed, and filter out the rare species; then, we will extract only the
observations for these more common species:

```{r, purl=FALSE}
## Extract the most common species_id
species_counts <- surveys_complete %>%
    count(species_id) %>%
    filter(n >= 50)

## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)
```

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
### Create the dataset for exporting:
##  Start by removing observations for which the `species_id`, `weight`,
##  `hindfoot_length`, or `sex` data are missing:
surveys_complete <- surveys %>%
    filter(species_id != "",        # remove missing species_id
           !is.na(weight),                 # remove missing weight
           !is.na(hindfoot_length),        # remove missing hindfoot_length
           sex != "")                      # remove missing sex

##  Now remove rare species in two steps. First, make a list of species which
##  appear at least 50 times in our dataset:
species_counts <- surveys_complete %>%
    count(species_id) %>%
    filter(n >= 50) %>%
    select(species_id)

##  Second, keep only those species:
surveys_complete <- surveys_complete %>%
    filter(species_id %in% species_counts$species_id)
```

To make sure that everyone has the same data set, check that `surveys_complete`
has `r nrow(surveys_complete)` rows and `r ncol(surveys_complete)` columns by
typing `dim(surveys_complete)`.

Now that our data set is ready, we can save it as a CSV file in our `data`
folder.

```{r, purl=FALSE, eval=FALSE}
write_csv(surveys_complete, file = "data/surveys_complete.csv")
```

```{r, purl=FALSE, eval=TRUE, echo=FALSE}
if (!dir.exists("data")) dir.create("data")
write_csv(surveys_complete, file = "data/surveys_complete.csv")
```