---
title: Scatterplots and Relationships
teaching: 20
exercises: 10
source: Rmd
---
  
::::::::::::::::::::::::::::::::::::::: objectives

- Create scatterplots to explore relationships between variables.
- Distinguish between aesthetic mappings and aesthetic settings.
- Add and modify layers in a ggplot.

::::::::::::::::::::::::::::::::::::::::::::::::::
  
:::::::::::::::::::::::::::::::::::::::: questions

- How can scatterplots be used to explore relationships between variables?
- How do aesthetic mappings and layers help reveal patterns in data?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-ggplot2, include=FALSE}
library(ggplot2)
```

We've seen how plots can be created quickly using the key components of a `ggplot` and we know that scatter plots can be used to explore relationships between variables in our dataset.

Let's take a closer look at the relationship between engine size (`displ`) and city fuel efficiency (`cty`). We will also explore how additional aesthetic mappings and layers can help reveal patterns in the data.

## Mapping and Setting Aesthetics

So far, we have mapped variables to the x- and y-axes. 

We can also control other visual properties, such as colour, size, and shape in two ways. 

We can **map** them to variables in our data, or **set** them to a fixed value that is applied to all observations. 

For example, we can **map** a colour to each `class` of vehicle.

```{r aes_map_colour, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points coloured by class."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty,
                     colour = class)) + 
  geom_point()
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Aesthetics - Colour trends

- What trends do the colours bring out in the data? 
- Are they what you expected?

:::::::::::::::  solution

- Vehicle classes form distinct clusters.
- SUVs and pickups generally have larger engines (higher displ) and lower city fuel efficiency (cty).
- Compact and subcompact cars generally have smaller engines and higher fuel efficiency.
- The overall negative relationship between engine size and fuel efficiency becomes clearer because we can see which vehicle classes - occupy different regions of the plot.
- The pattern is largely what we would expect: larger vehicle classes tend to be less fuel efficient than smaller vehicle classes.

:::::::::::::::::::::::::
  
::::::::::::::::::::::::::::::::::::::::::::::::::

In the above example we used a variable to control the colouring of the datapoints. 

We can also **set** a fixed value.

Lets modify the transparency of the points, using the `alpha` argument, which is especially helpful when you have a large amount of data which is very clustered.

```{r aes_set_alpha, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points coloured by class and shaped like triangles."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty,
                     colour = class)) + 
  geom_point(shape = 17)
```

Notice how all of the points are shaped like triangles, not just the ones of a particular class.

Finally, we can _map variables_ **and** _set values_ for aethestics for a **geom**. 

In this example, we will move the `colour` by `class` aesthetic to the specific `geom_point()`. 

This plot should look like the one above, we've only moved the colooring to the layer in order to make more complex visualisations.

```{r aes_set_map, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points coloured by class and shaped like triangles."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty)) + 
  geom_point(aes(colour = class), shape = 17)
```

Here the colour mapping only applies to the points because it was specified within `geom_point()`. 

This allows different layers to use different aesthetic mappings.

:::::::::::::::::::::::::::::::::::::::::  callout

## Aesthetic Mapping vs Setting

- Inside aes() = map an aesthetic to a variable in the data.
- Outside aes() = set an aesthetic to a fixed value.

| Goal    | Code |
| :---    | :--:|
| Colour by car class   | aes(colour = class)  |
| Make all points blue     | colour = "blue"  |
| Size by engine size | aes(size = displ)  |
| Make all points larger | size = 3  |

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

##  Mapping vs Setting

Modify the point layer in the previous example so that:

- all points are blue
- point size depends on `cty`

Hint: use `size` argument to change the point size.

:::::::::::::::  solution

Answers may vary.

Notice that the `city` argument is supplied inside the layer `aes()` function whereas the `colour` argument is supplied outside of `aes()`. This means that `colour` applies to all data points on the graph and is not related to a specific variable.

```{r ch-set, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points size according to the value of city mileage and coloured orange."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty)) +
  geom_point(aes(size = cty), colour = "orange")

```

:::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::::

Changing the way a dataset is displayed visually is useful for distinguishing patterns and extracting information. 

Taking it one step further, we can add more **geoms** to our plot to highlight relationships in the data.

## Layers

Layers in `ggplot2` are building blocks stacked on top of each other to create more and more complex plots.

Let's add a `geom_smooth()` layer to the plot. Recall we use the `+` symbol to stack the layers:
  
```{r cty-point, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points sized by city miledage and coloured orange. It includes a blue trendline with shaded confidence interval on top of the points."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty)) + 
  geom_point(aes(size = cty), colour = 'orange') +
  geom_smooth()
```

The `geom_smooth()` layer adds a trend line to the plot, making the overall relationship between engine size and fuel efficiency easier to see. In this case, it highlights that fuel efficiency tends to decrease as engine size increases.

It's important to note that each layer is drawn on top of the previous layer. In the plot above, the line has been drawn *on top of* the points. 

What if we wanted to see the points on top of the lines?

:::::::::::::::::::::::::::::::::::::::  challenge

## Layer Order Matters

Switch the order of the point and smooth layers from the previous example.
 
- What happened?

:::::::::::::::  solution

To demonstrate, rearrange the drawing order. 
  
The points now get drawn over the line! 

If we look closely the smooth line was drawn first, followed by the points.

```{r ch_order, fig.alt="Scatter plot of displacement vs city fuel efficiency with data points sized by city miledage and coloured orange. It includes a blue trendline with shaded confidence interval below the points."}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty)) + 
  geom_smooth() +
  geom_point(aes(size = cty), colour = 'orange')
```

:::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::::


In this episode, we used scatterplots to explore the relationship between engine size and fuel efficiency. 

We learned how aesthetic mappings can reveal additional patterns in the data and how aesthetic settings can be used to modify the appearance of a plot. 

Finally, we introduced layers and used `geom_smooth()` to highlight overall trends.

In the next episode, we will shift our focus from relationships between variables to the distribution of individual variables using density plots.

:::::::::::::::::::::::::::::::::::::::: keypoints

- Scatterplots are useful for exploring relationships between variables.
- Aesthetic mappings connect variables in the data to visual properties such as colour.
- Aesthetic settings apply the same visual property to all observations.
- `ggplot2` plots can be built up by adding layers with `+`.
- Different layers can have different aesthetic mappings.

::::::::::::::::::::::::::::::::::::::::::::::::::