---
title: relationships (20 min)
teaching: 20
exercises: 10
source: Rmd
---
  
::::::::::::::::::::::::::::::::::::::: objectives

- To be able to use ggplot2 to generate publication-quality graphics.
- To apply geometry, aesthetic mappings, and statistics layers to a ggplot plot.
- To manipulate the aesthetics of a plot using different colors, shapes, and lines.
- To improve data visualization through transforming scales and paneling by group.


::::::::::::::::::::::::::::::::::::::::::::::::::
  
:::::::::::::::::::::::::::::::::::::::: questions

- How can I create publication-quality graphics in R?
  
::::::::::::::::::::::::::::::::::::::::::::::::::
  
We've seen how plots can be created quickly using the key components of a `ggplot` and we know that scatter plots can be used to explore relationships between variables in our dataset.

Recall our displacement vs highway fuel efficiency plot with 'year' variable used to colour the dataset.


```{r ggplot-with-aes-geom-cty, eval=FALSE, fig.alt="Scatter plot of displacement vs highway fuel efficiency with data points coloured by year of manufacture."}}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy )) + 
  geom_point()
```

Let us now take a deeper dive into scatter plots by looking more closely at the relationship between fuel efficiency and displacement. We also want to add more aesthetic mappings to highlight these relationships and make our plot more easily understood.
  
## Layers
  
Using a scatter plot of points by itself is useful for visualizing how fuel efficiency changes as displacement increases. However, we can add a second **geom** to help make the relationship more obvious.

To add a trend line to the plot we use the `goem_smooth()` function:
  
```{r hwy-line, eval=FALSE, fig.alt=" scatter plot of displacement vs hwy fuel efficiency with data points and a smooth trendline.}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy, colour = year)) +
  geom_smooth()
```

Instead of adding a `geom_point` layer, we've added a `geom_line` layer.

However, the result doesn't look quite as we might have expected: it seems to be jumping around a lot in each continent. Let's try to separate the data by country, plotting one line for each country:

```{r lifeExp-line-by, eval=FALSE, fig.alt=""}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, group=country, color=continent)) +
  geom_line()
```

We've added the **group** *aesthetic*, which tells `ggplot` to draw a line for each
country.

But what if we want to visualize both lines and points on the plot? We can
add another layer to the plot:
  
  ```{r lifeExp-line-point}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, group=country, color=continent)) +
  geom_line() + geom_point()
```

It's important to note that each layer is drawn on top of the previous layer. In
this example, the points have been drawn *on top of* the lines. Here's a
demonstration:
  
  ```{r lifeExp-layer-example-1}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, group=country)) +
  geom_line(mapping = aes(color=continent)) + geom_point()
```

In this example, the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies
to the points. Now we can clearly see that the points are drawn on top of the
lines.

:::::::::::::::::::::::::::::::::::::::::  callout

## Tip: Setting an aesthetic to a value instead of a mapping

So far, we've seen how to use an aesthetic (such as **color**) as a *mapping* to a variable in the data. For example, when we use `geom_line(mapping = aes(color=continent))`, ggplot will give a different color to each continent. But what if we want to change the color of all lines to blue? You may think that `geom_line(mapping = aes(color="blue"))` should work, but it doesn't. Since we don't want to create a mapping to a specific variable, we can move the color specification outside of the `aes()` function, like this: `geom_line(color="blue")`.


::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 3

Switch the order of the point and line layers from the previous example. What
happened?

:::::::::::::::  solution

## Solution to challenge 3

The lines now get drawn over the points!

```{r ch3-sol, fig.alt="Scatter plot of life expectancy vs GDP per capita with a trend line summarising the relationship between variables. The plot illustrates the possibilities for styling visualisations in ggplot2 with data points enlarged, coloured orange, and displayed without transparency."}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, group=country)) +
 geom_point() + geom_line(mapping = aes(color=continent))
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Transformations and statistics

ggplot2 also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:
  
  ```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the x axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic. We can also modify the transparency of the
points, using the *alpha* function, which is especially helpful when you have
a large amount of data which is very clustered.

```{r axis-scale, fig.cap="Scatter plot of GDP vs life expectancy showing logarithmic x-axis data spread"}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10()
```

The `scale_x_log10` function applied a transformation to the coordinate system of the plot, so that each multiple of 10 is evenly spaced from left to right. For example, a GDP per capita of 1,000 is the same horizontal distance away from a value of 10,000 as the 10,000 value is from 100,000. This helps to visualize the spread of the data along the x-axis.

:::::::::::::::::::::::::::::::::::::::::  callout

## Tip Reminder: Setting an aesthetic to a value instead of a mapping

Notice that we used `geom_point(alpha = 0.5)`. As the previous tip mentioned, using a setting outside of the `aes()` function will cause this value to be used for all points, which is what we want in this case. But just like any other aesthetic setting, *alpha* can also be mapped to a variable in the data. For example, we can give a different transparency to each continent with `geom_point(mapping = aes(alpha = continent))`.


::::::::::::::::::::::::::::::::::::::::::::::::::

We can fit a simple relationship to the data by adding another layer,
`geom_smooth`:

```{r lm-fit, fig.alt="Scatter plot of life expectancy vs GDP per capita with a blue trend line summarising the relationship between variables, and gray shaded area indicating 95% confidence intervals for that trend line."}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10() + geom_smooth(method="lm")
```

We can make the line thicker by *setting* the **linewidth** aesthetic in the
`geom_smooth` layer:

```{r lm-fit2, fig.alt="Scatter plot of life expectancy vs GDP per capita with a trend line summarising the relationship between variables. The blue trend line is slightly thicker than in the previous plot."}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10() + geom_smooth(method="lm", linewidth=1.5)
```

There are two ways an *aesthetic* can be specified. Here we *set* the **linewidth** aesthetic by passing it as an argument to `geom_smooth` and it is applied the same to the whole `geom`. Previously in the lesson we've used the `aes` function to define a *mapping* between data variables and their visual representation.

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 4a

Modify the color and size of the points on the point layer in the previous
example.

Hint: do not use the `aes` function.

Hint: the equivalent of `linewidth` for points is `size`.

:::::::::::::::  solution

## Solution to challenge 4a

Here a possible solution:
  Notice that the `color` argument is supplied outside of the `aes()` function.
This means that it applies to all data points on the graph and is not related to
a specific variable.

```{r ch4a-sol, fig.alt="Scatter plot of life expectancy vs GDP per capita with a trend line summarising the relationship between variables. The plot illustrates the possibilities for styling visualisations in ggplot2 with data points enlarged, coloured orange, and displayed without transparency."}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(size=3, color="orange") + scale_x_log10() +
  geom_smooth(method="lm", linewidth=1.5)
```

:::::::::::::::::::::::::
  
  ::::::::::::::::::::::::::::::::::::::::::::::::::
  
  :::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 4b

Modify your solution to Challenge 4a so that the
points are now a different shape and are colored by continent with new
trendlines.  Hint: The color argument can be used inside the aesthetic.

:::::::::::::::  solution

## Solution to challenge 4b

Here is a possible solution:
  Notice that supplying the `color` argument inside the `aes()` functions enables you to
connect it to a certain variable. The `shape` argument, as you can see, modifies all
data points the same way (it is outside the `aes()` call) while the `color` argument which
is placed inside the `aes()` call modifies a point's color based on its continent value.

```{r ch4b-sol}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
 geom_point(size=3, shape=17) + scale_x_log10() +
 geom_smooth(method="lm", linewidth=1.5)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Multi-panel plots

Earlier we visualized the change in life expectancy over time across all
countries in one plot. Alternatively, we can split this out over multiple panels
by adding a layer of **facet** panels.

:::::::::::::::::::::::::::::::::::::::::  callout

## Tip

We start by making a subset of data including only countries located
in the Americas.  This includes 25 countries, which will begin to
clutter the plot.  Note that we apply a "theme" definition to rotate
the x-axis labels to maintain readability.  Nearly everything in
ggplot2 is customizable.


::::::::::::::::::::::::::::::::::::::::::::::::::

```{r facet}
americas <- gapminder[gapminder$continent == "Americas",]
ggplot(data = americas, mapping = aes(x = year, y = lifeExp)) +
  geom_line() +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 45))
```

The `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(~). This tells R to draw a panel for each unique value in the country column
of the gapminder dataset.

## Modifying text

To clean this plot up for a publication we need to change some of the text
elements. The x-axis is too cluttered, and the y axis should read
"Life expectancy", rather than the column name in the data frame.

We can do this by adding a couple of different layers. The **theme** layer
controls the axis text, and overall text size. Labels for the axes, plot
title and any legend can be set using the `labs` function. Legend titles
are set using the same names we used in the `aes` specification. Thus below
the color legend title is set using `color = "Continent"`, while the title
of a fill legend would be set using `fill = "MyTitle"`.

```{r theme}
ggplot(data = americas, mapping = aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  labs(
    x = "Year",              # x axis title
    y = "Life expectancy",   # y axis title
    title = "Figure 1",      # main title of figure
    color = "Continent"      # title of legend
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


This is a taste of what you can do with ggplot2. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc]. All RStudio cheat sheets are available from the [RStudio website][cheat_all].
Finally, if you have no idea how to change something, a quick Google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 5

Generate boxplots to compare life expectancy between the different continents during the available years.

Advanced:

- Rename y axis as Life Expectancy.
- Remove x axis labels.

:::::::::::::::  solution

## Solution to Challenge 5

Here a possible solution:
`xlab()` and `ylab()` set labels for the x and y axes, respectively
The axis title, text and ticks are attributes of the theme and must be modified within a `theme()` call.

```{r ch5-sol}
ggplot(data = gapminder, mapping = aes(x = continent, y = lifeExp, fill = continent)) +
 geom_boxplot() + facet_wrap(~year) +
 ylab("Life Expectancy") +
 theme(axis.title.x=element_blank(),
       axis.text.x = element_blank(),
       axis.ticks.x = element_blank())
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::