---
title: Data visualization with ggplot2
author: Data Carpentry contributors & Nathaniel Porter
teaching: 20
exercises: 10
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, purl=FALSE}
source("setup.R")
surveys_complete <- read_csv(file = "data_raw/portal_data_joined.csv", col_types = cols()) %>% drop_na(weight, hindfoot_length, sex) %>% group_by(species_id) %>% filter(n() >= 50) %>% ungroup()
```

:::::::::::::::::::::::::::::::::::::: questions 

- How do you make plots using R?
- How do you customize and modify plots?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Produce scatter plots and boxplots using `ggplot2`.
- Represent data variables with plot components.
- Modify the scales of plot components.
- Iteratively build and modify `ggplot2` plots by adding layers.
- Change the appearance of existing `ggplot2` plots using premade and customized themes.
- Describe what faceting is and apply faceting in `ggplot2`.
- Save plots as image files.

::::::::::::::::::::::::::::::::::::::::::::::::

We start by loading the required packages. **`ggplot2`** is included in the **`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

If not still in the workspace, load the data we saved in the previous lesson.

```{r load-data, eval=FALSE, purl=FALSE}
surveys_complete <- read_csv("data/surveys_complete.csv")
```

## Plotting with **`ggplot2`**

**`ggplot2`** is a plotting package that provides helpful commands to create complex plots
from data in a data frame. It provides a more programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data
change or if we decide to change from a bar plot to a scatterplot. This helps in
creating publication quality plots with minimal amounts of adjustments and
tweaking.

**`ggplot2`** refers to the name of the package itself. When using the package we use the
function **`ggplot()`** to generate the plots, and so references to using the function will
be referred to as **`ggplot()`** and the package as a whole as **`ggplot2`**

ggplot graphics are built layer by layer by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

1: use the `ggplot()` function and bind the plot to a specific data frame using the `data` argument.

```{r, eval=FALSE, purl=FALSE}
ggplot(data = surveys_complete)
```

2: define an aesthetic mapping (using the aesthetic (`aes`) function), by selecting the
variables to be plotted and specifying how to present them in the graph, e.g., as x/y
positions or characteristics such as size, shape, color, etc.

```{r, eval=FALSE, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
```

3: add 'geoms' – graphical representations of the data in the plot (points, lines, bars).
**`ggplot2`** offers many different geoms; we will use some common ones today, including:
  
  - `geom_point()` for scatter plots, dot plots, etc.
  - `geom_boxplot()` for, well, boxplots!
  - `geom_line()` for trend lines, time series, etc.

To add a geom to the plot use `+` operator. Because we have two continuous
variables, let's use `geom_point()` first:

```{r first-ggplot, purl=FALSE}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

The `+` in the **`ggplot2`** package is particularly useful because it allows
you to modify existing `ggplot` objects. This means you can easily set up plot
"templates" and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, eval=FALSE, purl=FALSE}
# Assign plot to a variable
surveys_plot <- ggplot(data = surveys_complete,
                       mapping = aes(x = weight, y = hindfoot_length))

# Draw the plot
surveys_plot +
  geom_point()
```

```{r, eval=FALSE, purl=TRUE, echo=FALSE}
## Create a ggplot and draw it.
surveys_plot <- ggplot(data = surveys_complete,
                       aes(x = weight, y = hindfoot_length))

surveys_plot +
  geom_point()
```

**Notes**

- Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x-
  and y-axis you set up in `aes()`.
- You can also specify aesthetics for a given geom independently of the
  aesthetics defined globally in the `ggplot()` function.
- The `+` sign used to add layers must be placed at the end of each line
  containing a layer. If, instead, the `+` sign is added in the line before the
  other layer, **`ggplot2`** will not add the new layer and will return an error
  message.
- You may notice that we sometimes reference 'ggplot2' and sometimes 'ggplot'.
  To clarify, 'ggplot2' is the name of the most recent version of the package.
  However, any time we call the function itself, it's just called 'ggplot'.
- The previous version of the **`ggplot2`** package, called **`ggplot`**,
  which also contained the `ggplot()` function is now unsupported and has
  been removed from CRAN in order to reduce accidental installations
  and further confusion.

```{r, eval=FALSE, purl=FALSE}
# This is the correct syntax for adding layers
surveys_plot +
  geom_point()

# This will not add the new layer and will return an error message
surveys_plot
  + geom_point()
```

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge (optional)

Scatter plots can be useful exploratory tools for small datasets. For data
sets with large numbers of observations, such as the `surveys_complete` data
set, overplotting of points can be a limitation of scatter plots. One strategy
for handling such settings is to use hexagonal binning of observations. The
plot space is tessellated into hexagons. Each hexagon is assigned a color
based on the number of observations that fall within its boundaries.  To use
hexagonal binning with **`ggplot2`**, first install the R package `hexbin`
from CRAN:

```{r, eval=FALSE}
install.packages("hexbin")
```

Then use the `geom_hex()` function:

```{r, eval=FALSE}
surveys_plot +
 geom_hex()
```

- What are the relative strengths and weaknesses of a hexagonal bin plot
  compared to a scatter plot? Examine the above scatter plot and compare it
  with the hexagonal bin plot that you created.

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r hexbin-challenge, echo=FALSE, eval=FALSE, purl=TRUE}
### Challenge with hexbin
##
## To use the hexagonal binning with **`ggplot2`**, first install the `hexbin`
## package from CRAN:

install.packages("hexbin")
library(hexbin)

## Then use the `geom_hex()` function:

surveys_plot +
    geom_hex()

## What are the relative strengths and weaknesses of a hexagonal bin
## plot compared to a scatter plot?
```

## Building your plots iteratively

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, purl=FALSE}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
    geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to avoid overplotting:

```{r adding-transparency, purl=FALSE}
ggplot(data = surveys_complete, aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.2)
```

We can also add colors for all the points:

```{r adding-colors, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.2, color = "blue")
```

### Adding another variable

Let's try coloring our points according to the sampling plot type (plot here refers to the physical area where rodents were sampled and has nothing to do with making graphs). Since we're now mapping a variable (`plot_type`) to a component of the ggplot2 plot (`color`), we need to put the argument inside `aes()`:

```{r color-by-species-1, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.2, aes(color = species_id))
```

:::::::::::::::::::::::::::::::::::::::  challenge

### Challenge

Use what you just learned to create a scatter plot of `weight` over
`species_id` with the plot types showing in different colors.
Is this a good way to show this type of data?

:::::::: solution

```{r scatter-challenge-answer, answer=TRUE, purl=FALSE}
ggplot(data = surveys_complete,
       mapping = aes(x = species_id, y = weight)) +
   geom_point(aes(color = plot_type))
```

:::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r scatter-challenge, echo=FALSE, eval=FALSE, purl=TRUE}
### Challenge with scatter plot:
##
##  Use what you just learned to create a scatter plot of `weight`
## over `species_id` with the plot types showing in different colors.
## Is this a good way to show this type of data?
```

## Changing scales

The default discrete color scale isn't always ideal: it isn't friendly to viewers with colorblindness and it doesn't translate well to grayscale. However, **`ggplot2`** comes with quite a few other color scales, including the fantastic `viridis` scales, which are designed to be colorblind and grayscale friendly. We can change scales by adding `scale_` functions to our plots:

```{r scale-viridis, warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) +
  geom_point(alpha = 0.2) +
  scale_color_viridis_d()
```

Scales don't just apply to colors- any plot component that you put inside `aes()` can be modified with `scale_` functions. Just as we modified the scale used to map `plot_type` to `color`, we can modify the way that `weight` is mapped to the `x` axis by using the `scale_x_log10()` function:

```{r scale-log, warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) +
  geom_point(alpha = 0.2) +
  scale_x_log10()
```

One nice thing about `ggplot` and the `tidyverse` in general is that groups of functions that do similar things are given similar names. Any function that modifies a `ggplot` scale starts with `scale_`, making it easier to search for the right function.

## Boxplot

We can use boxplots to visualize the distribution of weight within each species:

```{r boxplot, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot()
```

By adding points to the boxplot, we can have a better idea of the number of
measurements and of their distribution. Because the boxplot will show the outliers
by default these points will be plotted twice -- by `geom_boxplot` and
`geom_jitter`. To avoid this we must specify that no outliers should be added
to the boxplot by specifying `outlier.shape = NA`.

```{r boxplot-with-points, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, aes(color = plot_type))
```

Now our points are colored according to `plot_type`, but the boxplots are all the same color. One thing you might notice is that even with `alpha = 0.2`, the points obscure parts of the boxplot. This is because the `geom_point()` layer comes after the `geom_boxplot()` layer, which means the points are plotted on top of the boxes. To put the boxplots on top, we switch the order of the layers:

```{r reverse-layers, warning=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_jitter(aes(color = plot_type), alpha = 0.2) +
  geom_boxplot(outlier.shape = NA)
```

Now we have the opposite problem! The white `fill` of the boxplots completely obscures some of the points. To address this problem, we can remove the `fill` from the boxplots altogether, leaving only the black lines. To do this, we set `fill` to `NA`:

```{r fill-na, warning=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_jitter(aes(color = plot_type), alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, fill = NA)
```

Now we can see all the raw data and our boxplots on top.

## Changing themes

So far we've been changing the appearance of parts of our plot related to our data and the `geom_` functions, but we can also change many of the non-data components of our plot. 

At this point, we are pretty happy with the basic layout of our plot, so we can **assign** it to a plot to a named **object**. We do this using the **assignment arrow** `<-`. What we are doing here is taking the result of the code on the right side of the arrow, and assigning it to an object whose name is on the left side of the arrow.

We will create an object called `myplot`. If you run the name of the `ggplot2` object, it will show the plot, just like if you ran the code itself.

```{r}
myplot <- ggplot(data = surveys_complete, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_jitter(aes(color = plot_type), alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, fill = NA)

myplot
```

This process of assigning something to an **object** is not specific to `ggplot2`, but rather a general feature of R. We will be using it a lot in the rest of this lesson. We can now work with the `myplot` object as if it was a block of `ggplot2` code, which means we can use `+` to add new components to it.

We can change the overall appearance using `theme_` functions. Let's try a black-and-white theme by adding `theme_bw()` to our plot:

```{r theme-bw, warning=FALSE}
myplot + theme_bw()
```

As you can see, a number of parts of the plot have changed. `theme_` functions usually control many aspects of a plot's appearance all at once, for the sake of convenience. To individually change parts of a plot, we can use the `theme()` function, which can take many different arguments to change things about the text, grid lines, background color, and more. Let's try changing the size of the text on our axis titles. We can do this by specifying that the `axis.title` should be an `element_text()` with `size` set to 14.

```{r text-size, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14))
```

Another change we might want to make is to remove the vertical grid lines. Since our x axis is categorical, those grid lines aren't useful. To do this, inside `theme()`, we will change the `panel.grid.major.x` to an `element_blank()`.

```{r element-blank, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        panel.grid.major.x = element_blank())
```

Another useful change might be to remove the color legend, since that information is already on our x axis. For this one, we will set `legend.position` to "none". 

```{r legend-remove, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

```

::::::::::::::::::::::::::::: callout

Because there are so many possible arguments to the `theme()` function, it can sometimes be hard to find the right one. Here are some tips for figuring out how to modify a plot element:

- type out `theme()`, put your cursor between the parentheses, and hit <kbd>Tab</kbd> to bring up a list of arguments
    - you can scroll through the arguments, or start typing, which will shorten the list of potential matches
- like many things in the `tidyverse`, similar argument start with similar names
    - there are `axis`, `legend`, `panel`, `plot`, and `strip` arguments
- arguments have hierarchy
    - `text` controls all text in the whole plot
    - `axis.title` controls the text for the axis titles
    - `axis.title.x` controls the text for the x axis title

:::::::::::::::::::::::::::::

::::::::::::::::::::::::::::: callout

You may have noticed that we have used 3 different approaches to getting rid of something in `ggplot`: 

- `outlier.shape = NA` to remove the outliers from our boxplot
- `panel.grid.major.x = element_blank()` to remove the x grid lines
- `legend.position = "none"` to remove our legend

Why are there so many ways to do what seems like the same thing?? This is a common frustration when working with R, or with any programming language. There are a couple reasons for it:

1. Different people contribute to different packages and functions, and they may choose to do things differently.
2. Code may *appear* to be doing the same thing, when the details are actually quite different. The inner workings of `ggplot2` are actually quite complex, since it turns out making plots is a very complicated process! Because of this, things that seem the same (removing parts of a plot), may actually be operating on very different components or stages of the final plot.
3. Developing packages is a highly iterative process, and sometimes things change. However, changing too much stuff can make old code break. Let's say removing the legend was introduced as a feature of `ggplot2`, and then a lot of time passed before someone added the feature letting you remove outliers from `geom_boxplot()`. Changing the way you remove the legend, so that it's the same as the boxplot approach, could break all of the code written in the meantime, so developers may opt to keep the old approach in place.

:::::::::::::::::::::::::::::

## Changing labels

Our plot is really shaping up now. However, we probably want to make our axis titles nicer, and perhaps add a main title to the plot. We can do this using the `labs()` function:

```{r labels, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        legend.position = "none") +
  labs(title = "Rodent size by plot type",
       x = "Plot type",
       y = "Hindfoot length (mm)")
```

We removed our legend from this plot, but you can also change the titles of various legends using `labs()`. For example, `labs(color = "Plot type")` would change the title of a color scale legend to "Plot type".

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 3: Customizing a plot

Modify the previous plot by adding a descriptive subtitle. Increase the font size of the plot title and make it bold.

**Hint**: "bold" is referred to as a font "face"

:::::::::::::::::::::::: solution 

```{r customizing-challenge-answer, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), legend.position = "none",
        plot.title = element_text(face = "bold", size = 20)) +
  labs(title = "Rodent size by plot type",
       subtitle = "Long-term dataset from Portal, AZ",
       x = "Plot type",
       y = "Hindfoot length (mm)")
```

::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Faceting

One of the most powerful features of **`ggplot`** is the ability to quickly split a plot into multiple smaller plots based on a categorical variable, which is called **faceting**. 

So far we've mapped variables to the x axis, the y axis, and color, but trying to add a 4th variable becomes difficult. Changing the shape of a point might work, but only for very few categories, and even then, it can be hard to tell the differences between the shapes of small points.

Instead of cramming one more variable into a single plot, we will use the `facet_wrap()` function to generate a series of smaller plots, split out by `sex`. We also use `ncol` to specify that we want them arranged in a single column:

```{r facet-wrap, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        legend.position = "none", 
        panel.grid.major.x = element_blank()) +
  labs(title = "Rodent size by plot type",
       x = "Plot type",
       y = "Hindfoot length (mm)",
       color = "Plot type") +
  facet_wrap(vars(sex), ncol = 1)
```

::::::::::::::::::::::::::::: callout

Faceting comes in handy in many scenarios. It can be useful when:

- a categorical variable has too many levels to differentiate by color (such as a dataset with 20 countries)
- your data overlap heavily, obscuring categories
- you want to show more than 3 variables at once
- you want to see each category in isolation while allowing for general comparisons between categories

:::::::::::::::::::::::::::::::::::::

## Exporting plots

Once we are happy with our final plot, we can assign the whole thing to a new object, which we can call `finalplot`.

```{r assign-finalplot}
finalplot <- myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        legend.position = "none", 
        panel.grid.major.x = element_blank()) +
  labs(title = "Rodent size by plot type",
       x = "Plot type",
       y = "Hindfoot length (mm)",
       color = "Plot type") +
  facet_wrap(vars(sex), ncol = 1)
```

After this, we can run `ggsave()` to save our plot. The first argument we give is the path to the file we want to save, including the correct file extension. This code will make an image called `rodent_size_plots.jpg` in the `images/` folder of our current project. We are making a `.jpg`, but you can save `.pdf`, `.tiff`, and other file formats. Next, we tell it the name of the plot object we want to save. We can also specify things like the width and height of the plot in inches.

```{r save-plot, eval=FALSE}
ggsave(filename = "images/rodent_size_plots.jpg", plot = finalplot,
       height = 6, width = 8)
```

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 4: Make your own plot

Try making your own plot! You can run `str(complete_old)` or `?complete_old` to explore variables you might use in your new plot. Feel free to use variables we have already seen, or some we haven't explored yet.

Here are a couple ideas to get you started:

  - make a histogram of one of the numeric variables
  - try using a different color `scale_`
  - try changing the size of points or thickness of lines in a `geom`

::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: keypoints 

- the `ggplot()` function initiates a plot, and `geom_` functions add representations of your data
- use `aes()` when mapping a variable from the data to a part of the plot
- use `scale_` functions to modify the scales used to represent variables
- use premade `theme_` functions to broadly change appearance, and the `theme()` function to fine-tune
- start simple and build your plots iteratively

::::::::::::::::::::::::::::::::::::::::::::::::