---
title: Comparing Groups
teaching: 25
exercises: 20
source: Rmd
---
  
::::::::::::::::::::::::::::::::::::::: objectives

- Interpret boxplots using median, spread, and outliers
- Explain why boxplots are commonly used in statistical analysis
- Create boxplots to compare distributions across groups

::::::::::::::::::::::::::::::::::::::::::::::::::
  
:::::::::::::::::::::::::::::::::::::::: questions

- How can we summarise a distribution?
- What key information helps us understand a distribution quickly?
- How can we compare these summaries across groups?
  
::::::::::::::::::::::::::::::::::::::::::::::::::

In the previous episode, we used density plots to look at the distribution of a variable.

```{r dens-hwy, eval=FALSE, fig.alt="Density plot of highway fuel efficiency."}
ggplot(data = mpg, mapping = aes(x = hwy)) +
  geom_density()
```

These show the overall shape of the data. But sometimes we’re less interested in the full shape, and more in a quick summary. To do that, we can use a different type of plot.

## Introducing Boxplots

A boxplot is a visual summary of numerical data that displays its distribution, spread, and skewness in a compact way. 

We replace `geom_density()` with `geom_boxplot()` in our plot.

```{r box-hwy, eval=FALSE, fig.alt="Box plot of highway fuel efficiency."}
ggplot(data = mpg, mapping = aes(x = hwy)) +
  geom_boxplot()
```

Here we have single boxplot of highway fuel efficiency. This looks quite different to everything we've seen so far. Unlike density plots, this doesn’t show the full shape — it summarises the distribution.

What do you think this plot is showing?

### How to read a boxplot

Each boxplot shows:

- The **Median** is the line inside the box and indicates the “middle” value.
- The **Spread** or middle 50% is the box itself and shows where most values lie.
- **Whiskers** extend to typical lower and upper values.
- **Outliers** are individual points beyond the whiskers.


:::::::::::::::::::::::::::::::::::::::  challenge

## Interpreting the plot

- Where is the median fuel efficiency?
- How spread out are the middle values?
- Are there any outliers?

:::::::::::::::  solution

Answers may vary.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::  callout

## Density vs boxplot

**Density** plots show full shape, good for fewer groups.
**Boxplots** show summary, better for many groups.

Same underlying data, different level of detail.

::::::::::::::::::::::::::::::::::::::::::::::::::

## Do More With Box plots

Box plots can be used in many ways to extract more information from our data. They are especially common in statistical analysis and reporting.

So far, we’ve summarised a single distribution.

What if we want that same summary for each class of vehicle?

To get a boxplot for each `class`, we’re actually making two different changes here — first how we represent the data, and then how we organise it into groups. We must map:

- `class` → x (categorical group)
- `hwy` → y (numeric variable)

```{r box-hwy-class, eval=FALSE, fig.alt="Box plot of highway fuel efficiency."}
ggplot(data = mpg, mapping = aes(x=class, y = hwy)) +
  geom_boxplot()
```

Here, each box represents the distribution of one group. The summary values we learned above are calculated for each class of vehicle.

:::::::::::::::::::::::::::::::::::::::::  callout

## Mapping Change

geom_density: x = hwy
geom_boxplot (grouped): x = class, y = hwy

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Group Plot Interpretation

- Which vehicle class tends to have higher fuel efficiency?
- Which class has the widest spread?
- Which classes look similar or very different? 
- Are many outliers?

:::::::::::::::  solution

Answers may vary.

:::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Add fill to boxplot

Reuse what we already know and use `fill` to colour each class to help visually separate groups.

:::::::::::::::  solution

```{r box-hwy-class-colour, eval=FALSE, fig.alt="Box plot of highway fuel efficiency."}
ggplot(data = mpg, mapping = aes(x=class, y = hwy, fill = class)) +
  geom_boxplot()
```

:::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::::

In this episode, we moved from viewing the full shape of distributions to summarising them using boxplots. This allowed us to compare groups more clearly using a small number of key values.

Next, we’ll shift from comparing groups to understanding how observations make up a whole, using visualisations designed for proportions and composition.

:::::::::::::::::::::::::::::::::::::::: keypoints

- Boxplots summarise distributions using median, spread, and outliers
- Mapping x = group and y = numeric creates group comparisons
- Boxplots are useful when comparing many groups
- They trade detail (shape) for clarity (summary)

::::::::::::::::::::::::::::::::::::::::::::::::::

