---
title: Composition 
teaching: 25
exercises: 20
source: Rmd
---
  
::::::::::::::::::::::::::::::::::::::: objectives

- Create bar plots to show counts and proportions
- Map categorical variables to visualise composition
- Interpret plots as parts of a whole
- Recognise when bar plots are appropriate for composition

::::::::::::::::::::::::::::::::::::::::::::::::::
  
:::::::::::::::::::::::::::::::::::::::: questions

- How can we show how data are divided into parts?
- What does each category contribute to the whole?
  
::::::::::::::::::::::::::::::::::::::::::::::::::

In the previous episode, we used boxplots to summarise distributions and compare groups.

Now we’ll look at a different question. _How is our data made up?_

Instead of shape or summary, we’re interested in composition — how observations are divided into categories.

## Counts with bar plots

To start, let’s count how many observations fall into each vehicle `class`. As we learned with mapping, we map the group to the x-axis:

```{r bar-hwy, eval=FALSE, fig.alt="Bar plot showing the number of mpg observations for each class of vehicle ."}
ggplot(data = mpg, mapping = aes(x = class)) +
  geom_bar()
```

In this plot, `class` is a categorical variable and `geom_bar()` counts the number of observations in each category. The height of each bar shows how many there are. 

Notice that the y-axis shows counts. The taller the bar, the more observations belong to that class.


:::::::::::::::::::::::::::::::::::::::  challenge

## Interpreting the plot

- Which class appears most often?
- Which appears least often?

:::::::::::::::  solution

- suv
- 2seater

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


## Adding another variable

We can break this down further using `fill`, just like before. What if we wanted to know about the drive type (front-wheel, rear-wheel, 4-wheel) for the vehicles in each `class`. We use the `drv` variable. 

```{r bar-hwy-drv, eval=FALSE, fig.alt="Bar plot showing the number of mpg observations for each class of vehicle ."}
ggplot(data = mpg, mapping = aes(x = class, fill = drv)) +
  geom_bar()
```

Bars are now split into segments where each segment represents a level of `drv` or drive type.

The total height is still the count for each class so now we see composition within groups.

Notice that the overall height of each bar has not changed. We have simply divided each bar into drive-type categories.

:::::::::::::::::::::::::::::::::::::::  challenge

## Grouped Bar Plot Interpretation

- Which drive types are most common overall?
- Do some classes mostly use one drive type?
- Are some classes more mixed than others? 

:::::::::::::::  solution

Answers may vary.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Showing proportions

So far, we’ve shown counts.

Sometimes we care about proportions instead — how large each part is relative to the whole.

To show proportions instead of counts , add `position = "fill"` as an argument to `geom_bar()`:


```{r bar-hwy-prop, eval=FALSE, fig.alt="Bar plot showing the proportion of mpg observations for each class of vehicle and drive type."}
ggplot(data = mpg, mapping = aes(x = class, fill = drv)) +
  geom_bar(position = "fill")
```

At first glance, this plot looks similar to the previous one. However, notice that all bars now have the same height.

The y-axis has changed from counts to proportions. 

Instead of showing how many vehicles belong to each class, the bars now show the relative contribution of each drive type within a class.

This makes it easier to compare composition between classes, regardless of how many vehicles belong to each class.

:::::::::::::::::::::::::::::::::::::::  challenge

## Proportion Plot Interpretation

- What has changed on the y-axis?
- Why are all bars the same height?
- Which vehicle classes have the most similar composition?

:::::::::::::::  solution

Answers may vary.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

In this episode, we used bar plots to explore the composition of our data.

By mapping variables to fill, we broke counts into parts, and used proportions to compare how groups are made up.

So far, we've explored relationships, distributions, summaries, and composition. We've learned how to create several kinds of plots. Now let's make them easier for other people to understand.

:::::::::::::::::::::::::::::::::::::::: keypoints

- Bar plots show how data are divided into categories
- `geom_bar()` counts observations automatically
- Mapping `fill` shows composition within groups
- `position = "fill"` converts counts to proportions
- Composition focuses on parts of a whole, rather than shape or summary

::::::::::::::::::::::::::::::::::::::::::::::::::

