---
title: Communicating Results
teaching: 20
exercises: 15
source: Rmd
---
  
::::::::::::::::::::::::::::::::::::::: objectives

- Use `facet_wrap()` to create multiple panels from a grouped dataset
- Add informative titles and axis labels with `labs()`
- Improve readability by rotating crowded axis labels with `theme()`
- Refine an existing plot to better communicate a message

::::::::::::::::::::::::::::::::::::::::::::::::::
  
:::::::::::::::::::::::::::::::::::::::: questions

- How can we make plots easier to interpret?
- When is it helpful to split a plot into multiple panels?
- How can labels improve communication?
  
::::::::::::::::::::::::::::::::::::::::::::::::::

So far we've focused on creating plots to answer different questions:

- Relationships
- Distributions
- Group comparisons
- Composition

The final step is communication.

Even when a plot contains useful information, it may be difficult to read or interpret. ggplot2 provides tools that help us organise information and make plots clearer for our audience.

## Improving labels

Let's start with a plot we've seen before:

```{r bar-class, eval=FALSE, fig.alt="Bar plot of highway fuel efficiency for each class of vehicle."}
ggplot(data = mpg, mapping = aes(x = class)) +
  geom_bar()
```

The plot communicates the data, but the labels come directly from the dataset.

We can make the plot more informative using `labs()`:

```{r bar-class-labs, eval=FALSE, fig.alt="Bar plot of highway fuel efficiency for each class of vehicle."}
ggplot(data = mpg, mapping = aes(x = class)) +
  geom_bar() +
  labs(
    x = "Vehicle class",
    y = "Count",
    title = "Number of vehicles by class"
  )
```
  
### What changed?

- The x-axis now has a clearer description.
- The y-axis describes what is being measured.
- The plot has a descriptive title.

Good labels help readers understand a plot without needing to inspect the underlying code or dataset.

## Splitting plots with facets

Sometimes a single plot contains too much information.

In our composition episode, we examined vehicle classes and drive type:

```{r bar-hwy-comp, eval=FALSE, fig.alt="Bar plot showing the number of mpg observations for each class of vehicle ."}
ggplot(data = mpg, mapping = aes(x = class, fill = drv)) +
  geom_bar()
```

Instead of showing everything in one panel, we can create a separate panel for each drive type:

```{r bar-hwy-facet, eval=FALSE, fig.alt="Bar plot showing the number of mpg observations for each class of vehicle. Each drive type has its own panel."}
ggplot(data = mpg, mapping = aes(x = class)) +
  geom_bar() +
  facet_wrap(~ drv)
```

### What happened?

- A separate plot is created for each drive type.
- All panels use the same scale.
- Comparing patterns between groups becomes easier.

Faceting is useful when:

- Several groups overlap in a single plot
- You want to compare patterns between groups
- Colour alone is not enough to separate information

:::::::::::::::::::::::::::::::::::::::  challenge

## Facet Plot Interpretation

- Which drive type includes the most vehicle classes?
- Are some vehicle classes only found in one panel?
- Is the pattern easier to interpret than the single combined plot? 

:::::::::::::::  solution

Answers may vary.


:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


## Improving readability

Sometimes labels become crowded.

Consider the facet plot we just created. Notice that the x-axis labels are becoming crowded and difficult to read. How can we fix that?

We can rotate axis text to make it easier to read:

```{r bar-hwy-facet-lab, eval=FALSE, fig.alt="Bar plot showing the number of mpg observations for each class of vehicle. Each drive type has its own panel. Labels are rotated 45 degrees to improve readability."}
ggplot(data = mpg, mapping = aes(x = class)) +
  geom_bar() +
  facet_wrap(~ drv) +
  theme(axis.text.x = element_text(angle = 45))
```

Rotating labels prevents overlapping text, improves readability, and requires only a small modification.


:::::::::::::::::::::::::::::::::::::::  challenge

## Make a plot publication-ready

Choose one plot from a previous episode.

Improve it by:

- Adding a title
- Improving axis labels
- Rotating labels if necessary

:::::::::::::::  solution

Answers may vary.

```{r bar-hwy-facet-ch, eval=FALSE, fig.alt="Bar plot showing the number of mpg observations for each class of vehicle. Each drive type has its own panel."}
ggplot(data = mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "fill") +
  labs(
    title = "Drive type composition by vehicle class",
    x = "Vehicle class",
    y = "Proportion"
  ) +
  theme(axis.text.x = element_text(angle = 45))
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

In this episode, we focused on communicating information clearly using labels, facets, and simple formatting improvements. Throughout this workshop, we have used `ggplot2` to explore relationships, distributions, group comparisons, and composition. Clear communication helps ensure those visualisations can be understood and used by others.


:::::::::::::::::::::::::::::::::::::::: keypoints

- `labs()` adds informative titles and axis labels
- `facet_wrap()` creates a panel for each group
- Faceting can make complex plots easier to interpret
- Small theme adjustments can improve readability
- Effective visualisation includes clear communication, not just correct code

::::::::::::::::::::::::::::::::::::::::::::::::::

