---
title: "Extra Challenges"
teaching: 45
exercises: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 200, out.height = 600, out.width = 600, R.options = list(max.print = 100))
```

```{r load-tidyverse-package, message=FALSE}
library(tidyverse)
surveys <- read_csv("data/cleaned/surveys_complete_77_89.csv")
```

::::::::::::::::::::::::::::::::::::: challenge 

## Challenge: `ggplot2` syntax

There are some issues with these `ggplot2` examples. Can you figure out what is wrong with each one?

```{r, eval=FALSE}
ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length, color = "blue")) +
  geom_point()
```


:::::::::::::::::::::::: solution 

Our points don't actually turn out blue, because we defined the color inside of `aes()`. `aes()` is used for translating variables from the data into plot elements, like color. There is no variable in the data called "blue".

::::::::::::::::::::::::

```{r, eval=FALSE}
ggplot(data = surveys, 
       mapping = aes(x = "weight", y = "hindfoot_length")) +
  geom_point()
```

:::::::::::::::::::::::: solution 

Variable names inside `aes()` should not be wrapped in quotes.

::::::::::::::::::::::::

```{r, eval=FALSE}
ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length)) 
  + geom_point()
```

:::::::::::::::::::::::: solution 

When adding things like `geom_` or `scale_` functions to a `ggplot()`, you have to end a line with `+`, not begin a line with it.

::::::::::::::::::::::::

```{r, eval=FALSE}
ggplot(data = surveys, x = weight, y = hindfoot_length) +
  geom_point()
```

:::::::::::::::::::::::: solution 

When translating variables from the data, like `weight` and `hindfoot_length`, to elements of the plot, like `x` and `y`, you must put them inside `aes()`.

::::::::::::::::::::::::

```{r, eval=FALSE}
ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length, color = species_id)) +
  geom_point() +
  scale_color_continuous(type = "viridis")
```

:::::::::::::::::::::::: solution 

`species_id` is a categorical variable, but `scale_color_continuous()` supplies a continuous color scale. `scale_color_discrete()` would give a discrete/categorical scale.

::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::
